---
title: Checking a package
layout: default
output: bookdown::html_chapter
---

# Automated checking {#check}

`R CMD check` automatically checks your code for common problems. It's essential if you're planning on submitting to CRAN, but it's also useful even if you're not: it automatically detects many problems that you'd otherwise discover the hard way.

`R CMD check` will be frustrating the first time you run it - you'll discover many problems that you need to fix. The key to making `R CMD check` less frustrating is to run it more often. The sooner you find out about a problem, the easier it is to fix. The limit of this approach is to run `R CMD check` every time you make a change, and [Travis-CI](#travis) shows you how to do that if you use github.

## Workflow

`R CMD check` is so called, because that's the command you run from the console to check your package. I don't recommend calling `R CMD check` directly, instead run `devtools::check()`, or press Cmd + Shift + E in RStudio.

Compared to `R CMD check`, `devtools::check()`:

* Ensures that the documentation is up-to-date by first running 
  `devtools::document()`

* Checks a bundled package. This a best practice for checking because it starts
  from a clean slate: a package bundle doesn't contain any of the temporary 
  files that might have accumulated in your source package. For example,
  if you have compiled code, the source package will include compilation 
  artefacts like `.so` and `.o` files.

* Tries to be as close as possible to the checks that CRAN will run. This 
  includes setting a consistent locale, and setting a number of important flags 
  to `R CMD check`.

* Sets the `NOT_CRAN` environment variable which allows you to selectively
  skip tests on CRAN (see `testthat::skip_on_cran()` for more details).

`R CMD check` emits three types of message:

* `ERROR`: These are severe problems and you should fix regardless of whether
  or not you're submitting to CRAN.
  
* `WARNING`: These are likely problems. You must fix if you're planning to 
  submit to CRAN, and it's a good idea to look into them even if you're not.
  
* `NOTE`: Most of these are problems you should fix, but there are some false
  positives. If you're not submitting to CRAN, carefully consider each NOTE.
  If you are submitting to CRAN you should try really hard fix all NOTEs, even 
  if they are a false positive. It is possible to argue you way around
  them, but that requires human intervention. It's almost always better to 
  fix them so they don't appear, and no discussion is needed when you submit
  your package.

The checking workflow is simple, but tedious:

1. Run `devtools::check()`, or press Cmd + Shift + E.

2. Fix the first problem.

3. Repeat until there are no more problems.

## Checks

The following sections describe each individual check. I briefly describe what the check does and how to fix the most common problems. Use this list as a reference, when you have a problem with `R CMD check` and can't understand how to fix it. To make it easier to understand how the checks fit together, I've organised them into sections roughly corresponding to the chapters in this book. This means they will be in a somewhat different order to what you'll see when you run `check()`.

This list included every check run by R 3.1.1. If you're using a more recent version, you may want to consult the most recent version of this chapter online: <http://r-pkgs.had.co.nz/check.html>. If you have an error that this chapter doesn't help you with, please [email me](mailto:hadley@rstudio.com).

### Check metadata

`R CMD check` always starts by describing your current environment. I'm running R 3.1.1 on OS X with a UTF-8 charset:

* __Using log directory__ ‘/Users/hadley/Documents/web/httr.Rcheck’
* __Using R version__ 3.1.1 (2014-07-10)
* __Using platform__: x86_64-apple-darwin13.1.0 (64-bit)
* __Using session charset__: UTF-8

Next the description is parsed and the package version is printed. Here I'm checking httr version 0.5.0.9000 (you'll learn more about that weird version number in [versioning](#version)).

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3459 -->
* __Checking for file__ ‘httr/DESCRIPTION’
* __This is package__ ‘httr’ __version__ ‘0.5.0.9000’

### Package structure

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L257 -->
* __Checking package directory__. The directory you're checking must exist - 
  `devtools::check()` protects you against this problem.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3720 -->
* __Checking if this is a source package__. This ensures that you're checking
  a source package, not an binary or installed package. Using 
  `devtools::check()` ensures that you will never have a problem here.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3015 -->
* __Checking for executable files__. You should not have an executable files in
  your package: they're not portable, the source code is not available, and 
  they are a security risk. Delete any executables files from your package. 
  (If you're not submitting to CRAN, and want to silence this warning, you can 
  list executable files in the `DESCRIPTION` field `BinaryFiles`.)

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3103 -->
* __Checking for hidden files and directories__. On linux and OS X, files with
  a name starting with `.` are hidden by default, and you've probably included
  them in your package by mistake. Either delete them, or if they are important, 
  use `.Rbuildignore` to remove them from the package bundle. R automatically 
  removes some common directories like `.git` and `.svn`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L381 -->
* __Checking for portable file names__.  Since R packages must work on windows, 
  linux and OS X, only use file names that work on all platforms. The easiest 
  way to do this is to stick to letters, numbers and underscores and dashes. 
  Avoid non-English letters and spaces. Rename any files listed in this check. 
  
<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L541 -->
* __Checking for sufficient/correct file permissions__. If you can't read a 
  file, you can't check it. This check detects the unlikely occurence that 
  you have files in the package that you don't have permission to read. Fix this
  problem by fixing the file permissions.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3141 -->
* __Checking whether package ‘XYZ’ can be installed__. `R CMD check` runs
  `R CMD install` to make sure that it's possible to install your package. If 
  this fails, you should run `devtools::install()` or RStudio's Build & Reload
  and debug any problems before continuing.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3459 -->
* __Checking installed package size__. It's easy to accidentally include large
  files that blow up the size of your package. This check ensures that the
  whole package less than 5 MB and each subdirectory is be less than 1 MB. 
  If you see this message, check that you haven't accidentally included a 
  large file. 
  
    If submitting to CRAN, you'll need to justify the size of your package. 
    First, make sure the package is as small as it possibly can be: try 
    recompressing the data, [data CRAN notes](#data-cran); and minimising 
    vignettes, [vignette CRAN notes](#vignette-cran) for more details.
    If it's still too large, consider moving data into its own package.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L746 -->
* __Checking top-level files__. Only specified files and directories are 
  allowed at the top level of the package (e.g. `DESCRIPTION`, `R/`, `src/`). 
  To include other files, you have two choices:
  
    * If they don't need to be installed (i.e. they're only used in the 
      source package): add to `.Rbuildignore` with 
      `devtools::add_build_ignore()`.
      
    * If they need to be installed: put `inst/`. They'll be moved to the 
      top level when the package is installed.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L924 -->
* __Checking package subdirectories__. 

    * Don't include any sub-directories. These are usually removed 
      automatically by `R CMD build` so you shouldn't see this error.
      
    * The case of files and directories is important. All sub-directories 
      should be lower-case, except for `R/`. A citation file, if present, 
      should be in `inst/CITATION`. Rename as needed.
      
    * The contents of `inst/` shouldn't clash with top-level contents 
      of the package. If they do, rename your files/directories.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L859 -->
* __Checking for left-over files__. Remove any files listed here. They've 
  somehow ended up in your package by accident.

### Description

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L597 -->
* __Checking DESCRIPTION meta-information__. 

    * The `DESCRIPTION` must be valid. You are unlikely to see this error,
      because `devtools::load_all` runs the same check each time your re-load
      the package.
    
    * If you use any non-ASCII characters in the DESCRIPTION, you must also
      specify an encoding. There are only three encodings that work on all
      platforms: latin1, latin2 and UTF-8. I strongly recommend UTF-8.
    
    * The `License` must refer to either a known license (a complete list can 
      be found at <https://svn.r-project.org/R/trunk/share/licenses/license.db>),
      or it must use `file LICENSE` and that file must exist. Errors here are 
      most likely to be typos.
      
    * You should either provide `Authors@R` or `Authors` and `Maintainer`. 
      If you're specified both, you'll get an error, so remove the one you
      didn't want.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3683
tools:::.check_package_depends -->
* __Checking package dependencies__. 

    * All packages listed in `Depends`, `Imports` and `LinkingTo` must be 
      installed, and their version requirements must be met, otherwise your 
      package can't be checked. An easy way to install any missing or outdated
      dependencies is to run `devtools::install_deps(deps = T)`.
`  
    * Packages listed in `Suggest` must be installed, unless you've set 
      the environment variable `_R_CHECK_FORCE_SUGGESTS_` to a false value.
      This allows is useful if some of the suggested packages are only available
      on certain platforms.
    
    * R packages can not have a cycle of dependencies: i.e. if package A 
      requires B, then B can not require A (otherwise which one would you load
      first?). If you see this error, you'll need to rethink the design of 
      your package. One easy fix is to make the package suggested.

    * Any packages used in the `NAMESPACE` must be listed in one of 
      `Imports` or `Depends`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3561 
tools:::.check_package_CRAN_incoming
-->
* __Checking CRAN incoming feasibility__. These checks only apply if you're 
  submitting to CRAN.
  
    * Your package can not use the same name as existing package (unless you're 
      the maintainer of that package), and the version number must be higher 
      than the current CRAN version.
        
    * If the maintainer of the package has changed (even if it's just a change 
      in email address), the new maintainer should submit to CRAN, and the
      old maintainer should send a confirmation email.

    * You must use a standard open source license, as listed in 
      <https://svn.r-project.org/R/trunk/share/licenses/license.db>.
    
    * The title and description must be spelled correctly. The title of
      the package must be in title case. Neither title not description
      should include either the name of your package or the word "package".
    
    * You'll get a note if it's a new package for CRAN - this reminds the
      CRAN maintainers to do some extra manual checks.
      
    * Avoid submitting multiple versions of the same package in a short
      period of time. CRAN prefers at most one submission per month. If you
      need to fix a major bug, be apologetic.

### Namespace

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L4433 -->
* __Checking if there is a namespace__. You must have a `NAMESPACE` file.
  Roxygen2 will create this for you as described in [namepsaces](#namespace).

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L3620 -->
* __Checking package namespace information__. The `NAMESPACE` should be 
  parseable by `parseNamespaceFile()` and valid. If this check fails, it's
  a bug in roxygen2.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2209 -->
* __Checking whether the package can be loaded__. R loads your package with
  `library()`. Failure here typically indicates a problem with 
  `.onLoad()` or `.onAttach()`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2234 -->
* __Checking whether the package can be loaded with stated dependencies__.
  Runs `library(pkg)` with `R_DEFAULT_PACKAGES=NULL`, so the search path 
  is empty. Failure here typically indicates that you're missing a dependency.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2247 -->
* __Checking whether the package can be unloaded cleanly__. Loads with 
  `library()` and then `detach()`es. If this fails, check `.onUnload()` and 
  `.onDetach()`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2264 -->
* __Checking whether the namespace can be loaded with stated dependencies__.
  Run `loadNamespace(pkg)` with `R_DEFAULT_PACKAGES=NULL`. Failure usually 
  indicates a problem with the namespace.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2279 -->
* __Checking whether the namespace can be unloaded cleanly__.
  Runs `loadNamespace("pkg"); unloadNamespace("pkg")`. Check `.onUnload()` for 
  problems.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2300 -->
* __Checking loading without being on the library search path__.
  Calls `library(x, lib.loc = ...)`. Failure here indicates that you are 
  making a false assumption in `.onLoad()` or `.onAttach()`.

### R code

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1226 -->
* __Checking R files for non-ASCII characters__. For maximum portability (i.e.
  so people can use your package on Windows) you should avoid using non-ASCII
  characters in R files. It's ok to use them in comments, but object names 
  shouldn't use them, and in strings you should use unicode escapes. See 
  [R/ CRAN notes](#r-cran) for more deatils.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1258 -->
* __Checking R files for syntax errors__. Obviously your R code must be valid.
  You're unlikely to see this error if you're been regularly using 
  `devtools::load_all()`.

<!--- tools:::.check_packages_used --->
* __Checking dependencies in R code__. Any errors here indicate that you've 
  forgotten to declare a needed package in the `DESCRIPTION`. Remember that you 
  should never use `require()` or `library()` inside a package - see 
  [namespace imports](#imports) for more details on best practices.

<!--- tools::checkS3methods --->
* __Checking S3 generic/method consistency__. S3 methods must have compatible 
  method signatures to their generic. This means that the method must have the 
  same arguments as its generic, with one exception. If the generic includes 
  `...` the method can have additional arguments. A common cause of this error 
  is defining print methods, because the `print()` generic contains`...`.
  
    ```{r}
    print
    
    # BAD
    print.my_class <- function(x) cat("Hi")
    
    # GOOD
    print.my_class <- function(x, ...) cat("Hi")
    ```

<!-- tools::checkReplaceFuns -->
* __Checking replacement functions__. Replacement functions (e.g. functions that
  are called like `foo(x) <- value`, must have `value` as the last argument.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1373
     tools:::.check_package_code_shlib
     tools:::.check_package_code_startup_functions
     tools:::.check_package_code_unload_functions
     tools:::.check_package_code_tampers
     tools:::.check_code_usage_in_package
     tools:::.check_dotInternal
     tools:::.check_package_code_assign_to_globalenv
     tools:::.check_package_code_attach
     tools:::.check_package_code_data_into_globalenv
     tools:::.check_depdef
     -->
     
* __Checking R code for possible problems__. This is a catch all for
  a wide range of problems:

    * Calls to `library.dynam()` (and `library.dynam.unload()`) should look
      like `library.dynam("name")`, not `library.dynam("name.dll")`. Remove
      the extension to fix this error.
      
    * Put `library.dynam()` in `.onLoad()`, not `.onAttach()`;
      put `packageStartupMessage()` in `.onAttach()`, not `.onLoad()`.
      Put `library.dynam.unload()` in `.onUnload()`.
  
    * Don't use `unlockBinding()` or `assignInNamespace()` to modify objects
      that don't belong to you. 
      
    * `codetools::checkUsagePackage()` is called to check that your functions
      don't use variables that don't exist. This sometimes raises false 
      positives with functions that use non-standard evaluation (NSE),
      like `subset()` or `with()`. Generally, I think you should avoid NSE in 
      package functions, and hence avoid this NOTE, but if there is no 
      standard evaluation version, see `?globalVariables` to suppress the
      NOTE.
      
    * You are not allowed to use `.Internal()` in a package. Either call 
      the R wrapper function, or write your own C function. (If you copy and 
      paste C function from base R, make sure to maintain the copyright notice,
      and your package will need to use a GPL-2 compatible license.)

    * Don't use `assign()` to modify objects in the global environment. If 
      you need to maintain state across function calls, create your own
      environment with `e <- new.env(parent = emptyenv())` and set and 
      get values in it.
  
    * Don't use `attach()` in your code. Instead refer to variables 
      explicitly.
  
    * Don't use `data()` without specifying the `envir` argument. Otherwise
      the data will be loaded in the global environment.
      
    * Don't use deprecated or defunct functions. Update your code to use 
      the latest versions.
  
    * You must use `TRUE` and `FALSE` in your code and examples, not `T` and 
      `F`.
    
### Data

<!-- 
https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1782 
tools:::.check_package_datasets
tools:::.check_package_compact_datasets
tools:::.check_package_compact_sysdata
-->
* __Checking contents of 'data' directory__.

    * The data directory can only contain file types described in 
      [exported data](#data-data).
      
    * Data files can only contain non-ASCII character if the encoding is 
      not correctly set. This usually shouldn't be a problem if you're saving
      `.Rdata` files. If you do see this error, look at the `Encoding()` of
      each column in the data frame, and ensure none are "unknown" (you'll 
      typically need to fix this somewhere in the import process).
      
    * If you've compressed a data file with `bzip2` or `xz` you need to declare
      at least `Depends: R (>= 2.10)` in your `DESCRIPTION`.
      
    * If you've picked a poor compression for your data, re-run with the 
      best suggested.

### Documentation

You can run the most common of these outside `R CMD check` with `devtools::check_doc()` (which automatically calls `devtools::document()` for you). If you have documentation problems, it's best to iterate quickly with `check_doc()`. 

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1527 -->
* __Checking Rd files__. This checks that all `man/*.Rd` use the correct
  Rd syntax. If this fails for you, it indicates a bug in roxygen2. Please 
  let me know.

<!-- tools:::.check_Rd_metadata -->
* __Checking Rd metadata__. Checks that names and aliases are unique. If you
  encounter this problem you've accidentally used the same `@name` or `@aliases`
  in multiple places; make sure they're unique.

<!-- tools:::.check_Rd_line_widths -->
* __Checking Rd line widths__.  Lines in Rd files must be less than 90 
  characters wide. This is unlikely to occur if you wrap your R code,
  and hence roxygen comments to 80 characters. For very long urls, use a 
  link-shortening service like [bit.ly](http://bit.ly).

<!-- tools:::.check_Rd_xrefs -->
* __Checking Rd cross-references__. Errors here usually represent typos.
  Also recall the syntax for linking to functions in other packages: 
  `\link[package]{function}`. Sometimes I accidentally flip `\code{}` and
  `\link{}`: `\link{\code{function}}` will not work.

<!-- tools::undoc -->
* __Checking for missing documentation entries__. All exported objects must 
  be documented. All paramters of functions must be documented. See 
  `?tools::undoc` for more details.

<!-- tools::codoc, tools::codocData, tools::codocClasses -->
* __Checking for code/documentation mismatches__. This make sure that 
  documentation matches the code. This should never fail because you're using
  roxygen2 which automatically keeps them in sync.

<!-- tools::checkDocFiles, tools::checkDocStyle -->
* __Checking Rd `\usage` sections__.  Checks that all arguments in funtions
  in usage section are documented (and all `@params` document an existing
  argument). S3 and S4 methods need to use special markup ``\S3method{}`,
  `\S4method{}` in the Rd file - roxygen2 should generate this for you
  automatically.

<!-- tools:::.check_Rd_contents -->
* __Checking Rd contents__. This checks for autogenerated contect made by
  `package.skeleton()`. Since you're not using `package.skeleton()` you should
  never have a problem here.

<!-- tools:::.check_packages_used_in_examples -->
* __Checking for unstated dependencies in examples__. If you use a package 
  only for an example, make sure it's listed in the `Suggests` field. Before
  running example code that depends on it, test to see if it's available with
  `requireNamespace("pkg", quietly = TRUE)`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2317 -->
* __Checking examples__. Every documentation example must run without errors,
  and must not take too long. Exclude failing or slow tests with `\donttest{}`.
  See [documenting functions](#man-functions) for more details.
  
    A common cause of problems is relying on internal functions - examples can 
    only use exported functions.
    
    Examples are one of the last checks run, so first them can be painful if
    you have to run `devtools::check()` each time. Instead, use 
    `devtools::run_examples()`: it only checks the examples, and has an optional 
    parameter which tells it which function to start at - that way once you've 
    discovered an error, you can rerun from just that file, not all the 
    files that lead up to.
    
    Don't open new graphics devices.
    
    Don't use more than two cores.
    
    Individual examples shouldn't take more than 5s.
    
<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2925 -->
* __Checking PDF version of manual__. Occassionally you'll get an error when 
  building the PDF manual. This is usually because the pdf is built by latex and 
  you've forgotten to escape something. Debugging this is painful - your best 
  bet is to look up the latex logs and combined tex file and work back from 
  their to `.Rd` files then back to a roxygen comment. I consider any such
  failure to be a bug in roxygen2, so please let me know.

### Demos

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L877 -->
* __Checking index information__. If you've written demos, each demo must be 
  listed in `demos/00Index`. NB: the format of this file is 
  `demo-name-with-out-extension  Demo description`.

### Compiled code

<!-- tools::checkFF -->
* __Checking foreign function calls__. `.Call()`, `.C()`, `.Fortran()`, 
  `.External()` must always be called either with a `NativeSymbolInfo` object
  (as created with `@useDynLib`) or use the `.package` argument.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2065 -->
* __Checking line endings in C/C++/Fortran sources/headers__. Always
  use LF as a line ending.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2084 -->
* __Checking line endings in Makefiles__. As above.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2120 -->
* __Checking for portable use of `$(BLAS_LIBS)` and `$(LAPACK_LIBS)`__.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2159 
 tools:::check_compiled_code
 -->
* __Checking compiled code__.  Checks that you're not using any C functions
  that you shouldn't.

### Tests

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2514 -->
* __Checking for unstated dependencies in tests__. Every package used by tests 
  must be included in the dependencies.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2526 -->
* __Checking tests__.  Each file in `tests/` is run. If you've followed the 
  instructions in [testing](#tests) you'll have at least one file: 
  `testthat.R`. If your unit tests don't pass fix them!

### Vignettes

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L718 -->
* __Checking ‘build’ directory__. Used to track vignette builds. I'm not
  sure how this could fail unless you've accidentally `.Rbuildignore`d a file.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1846 -->
* __Checking installed files from ‘inst/doc’__.  Don't put files in `inst/doc` -
  vignettes now live in `vignettes/`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L1941 -->
* __Checking files in ‘vignettes’__. 
  Problems here are usually straightforward - you've included files that are
  already included in R (like `jss.cls`, `jss.bst`, or `Sweave.sty`), or 
  you have leftover latex compilation files. Delete these files.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2008 -->
* __Checking for sizes of PDF files under 'inst/doc'__. If you're making PDF
  vignettes, you can make them as small as possible by running 
  `tools::compactPDF()`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2619 -->
* __Checking for unstated dependencies in vignettes__. As with tests, every
  package that you use in a vignette must be listed in the `DESCRIPTION`. 
  If a package is used only for a vignette, and not elsewhere, make sure 
  it's listed in `Suggests`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2629 -->
* __Checking package vignettes in ‘inst/doc’__. This checks that every source
  vignette (i.e. `.Rmd`) has a built equivalent (i.e. `.html`) in `inst/doc`. 
  This shouldn't fail if you've used the standard process outlined in 
  [vignettes](#vignettes). If there is a problem, start by checking your
  `.Rbuildignore`.
  
<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2763 -->
* __Checking running R code from vignettes__. The R code from each vignette 
  is run. If you want to deliberately execute errors (to show the user what 
  failure looks like), make sure the chunk has `error = TRUE, purl = FALSE`.

<!-- https://github.com/wch/r-source/blob/trunk/src/library/tools/R/check.R#L2856 -->
* __Checking re-building of vignette outputs__. Each vignette is re-knit to
  make sure that the output corresponds to the input. Again, this shouldn't
  fail in normal circumstances.
  
To run vignettes, the package first must be installed. That means `check()`:

* Builds the package.
* Installs the package without vignettes.
* Builds all the vignettes.
* Re-installs the package with vignettes.

If you have a lot of compiled code, this can be rather slow. You may want to add `--no-build-vignettes` to the commands list in "Build Source Packages" field in the project options:

```{r, echo = FALSE}
bookdown::embed_png("screenshots/project-options-build.png", dpi = 220)
```

## Checking after every commit with Travis {#travis}

If you use [git and github](#git), I recommend exploring the very useful [Travis](https://travis-ci.org/) continuous integration service. A continuous integration service runs code of your choosing every time you push a commit to github. 

To use travis:

1.  Run `devtools::use_travis()` to set up a basic `.travis.yml` config file. The 
    defaults are usually adequate for simple packages.

1.  Navigate to your [account](https://travis-ci.org/profile) and enable
    travis for the repo you want to test on.

1.  Commit and push to github.

1.  Wait a few minutes to see the results in your email.

Now you're set up so that `R CMD check` will be run every time you push a commit to github, and every time some one does a pull request.  Any time a failure occurs you'll find out about it right away. I find this really useful - it also encourages me to run `R CMD check` more frequently locally, because I know if it fails I'll find out about a few minutes later, often once I've moved on to a new challenge.

### Customising your config

Common configuration:

* Fail not only on errors, but also on warnings
* Install binary packages
* Install github packages

Since you can use Travis to run arbitrary code you can use it to implement arbitrary push-to-deploy strategies. For example, it's relatively easy to deploy code to S3 - this makes it easy (for example) to update a documentation website.
