---
layout: default
title: Data
output: bookdown::html_chapter
---

# External data {#data}

It's often useful to include data in a package. If you're releasing the package to a broad audience, an interesting dataset is useful for building compelling examples. Or you might be interested primarily in the data, and a package is a good way to distributed it along with documentation.

There are four ways to include data in your package, depending on what you want to do with it and who should be able to use it:

* If you want to store parsed data, and make it available to the user, put it
  in `data/`. This is the best place to put example datasets.
  
* If you want to store parsed data, but not make it available to the user,
  put it all `R/sysdata.rda`. This is the best place to put data that your
  functions need.

* If you want to store raw data, put it in `inst/extdata`.

* If the data is only used by a vignette, it in the `vignettes/`.

Each possible location is described in more detail below.

## `data/`

The most common location for package data is (surprise!) `data/`. Each file in this directory should be a `.rda` file created by `save()` containing a single object (with the same name as the file). (There are other ways to store data, but this is the best - it takes up the least space and is the fastest to load.) The easiest way to adhere to these rules is to use `devtools`::use_data()`:

```{r, eval = FALSE}
x <- sample(1000)
devtools::use_data(x, mtcars)
```

If the `DESCRIPTION` contains `LazyData: true`, then datasets will be lazily loaded. This means that they don't occupy any memory until you use them. The following example shows the memory usage before and after loading the nycflights13 package. The memory usage doesn't change until you inspect the flights dataset stored inside the package. 

```{r}
pryr::mem_used()
library(nycflights13)
pryr::mem_used()

invisible(flights)
pryr::mem_used()
```

I recommend that you always include `LazyData: true` in your `DESCRIPTION`. `devtools::create()` does this for you.

Often you'll create the data files in `data/` from raw data gathered from elsewhere. I highly recommend taking the time to make this code fully reproducible so that you can re-run whenever you need to update the data. I suggest that you put this code in `data-raw/`. You'll need to add to the `.Rbuildignore` so it doesn't get included in the distributed packgage. Do this automatically with:

```{r, eval = FALSE}
devtools::use_data_raw()
```

You can see this approach in practice in some of my recent data packages. I've been creating these packages because the data will change rarely, and then multiple packages can use them for examples:

* [babynames](https://github.com/hadley/babynames)
* [fueleconomy](https://github.com/hadley/fueleconomy)
* [nasaweather](https://github.com/hadley/nasaweather) 
* [nycflights13](https://github.com/hadley/nycflights13)
* [usdanutrients](https://github.com/hadley/usdanutrients)

### Documenting datasets {#documenting-data}

Objects in `data/` are always effectively exported (they use a slightly different mechanism to the `NAMESPACE` but the details are not important). This means that the must be documented. Documenting data is like documenting a function with a few minor differences. Instead of documenting the data directly, you document the name of the dataset. For example, the roxygen2 block used to document the diamonds data in ggplot2 looks something like this:

```{r, eval = FALSE}
#' Prices of 50,000 round cut diamonds.
#'
#' A dataset containing the prices and other attributes of almost 54,000
#' diamonds.
#'
#' @format A data frame with 53940 rows and 10 variables:
#' \describe{
#'   \item{price}{price, in US dollars}
#'   \item{carat}{weight of the diamond, in carats}
#'   ...
#' }
#' @source \url{http://www.diamondse.info/}
"diamonds"
```

There are two additional tags that are important for documenting datasets:

* `@format`, which gives an overview of the dataset. For data frames, you 
  should include a definition list that describes each variable. It's usually
  a good idea to describe units here.
  
* `@source` where you got the data form, often a `\url{}`.

Never `@export` a data set.

## `R/sysdata.rda`

Sometimes functions need pre-computed data tables. If you put these in `data/` they'll also be available to package users, which is not appropriate. Instead, you can save them in `R/sysdata.rda`. For example, two coloured related packages, [munsell](https://github.com/cwickham/munsell) and [dichromat](http://cran.r-project.org/web/packages/dichromat/index.html), use `R/sysdata.rda` to store large tables of colour data.

You can use `devtools::use_data()` to create this file with the argument `internal = TRUE`:

```{r, eval = FALSE}
x <- sample(1000)
devtools::use_data(x, mtcars, internal = TRUE)
```

Again, it's a good idea to make the code used to generate this file reproducible. Put it in `data-raw/`.

Objects in `R/sysdata.rda` are not exported (and shouldn't be), so don't need to be documented. They're only available inside your package.

## `inst/`

If you want to show examples of loading/parsing raw data, put the original files in `inst/extdata`. When the package is installed, all files in `inst/` are moved into the top-level directory (so they can't have names like `R/` or `DESCRIPTION`). To refer to files in `inst/extdata` (whether installed or not), use `system.file()`:

```{r}
system.file("include", "Rcpp.h", package = "Rcpp")
```

Beware: if the file does not exist, `system.file()` does not return an error - it just returns the empty string:

```{r}
system.file("include", "Rcp.h", package = "Rcpp")
```

## `vignettes/`

If you need data for a vignette, it's fine to just include it in the vignettes directory. Refer to it with a relative path. For example, if you're vignette is `vignette/my-vignette.Rmd` and your data is `vignette/my-data.csv`, then in `my-vignette.Rmd` you'd do `read.csv("my-data.csv")`. (You'll quickly discovered if you did this incorrectly because the vignette will fail to knit.)

The data will be included in the package bundle, but not the installed package. This makes it difficult to re-run the code in the vignette. You can force R to include the data file in the installed package by including the path in `vignettes/.install_extras`. Like `.Rbuildignore` this file uses Perl regular expressions, so it's best to use a helper function:

```{r, eval = FALSE}
devtools::use_vignette_extra("vignettes/mtcars.csv")
```

## CRAN notes {#data-cran}

If you are submitting your package to CRAN, you will need to make sure that the data has been optimally compressed (and it's useful to do so even if you're not submitting). Run `tools::checkRdaFiles()` to determine the best compression for each file. If you've lost the code for recreating the files, you can use `tools::resaveRdaFiles()` to save as best format, but it's better to modify the original `save()` code. Generally, package data should be smaller than a megabyte - if it's larger you'll need to argue for an exemption. This is usually easier to do if the data is its own package, and it won't be updated frequently.

