---
title: Package basics
layout: default
output: bookdown::html_chapter
---

# Package metadata {#description}

The job of the `DESCRIPTION` file is to store important metadata about your package. When you first start writing packages, you'll mostly use these metadata to record what packages are needed to run your package. However, as time goes by and you start sharing your package with others, the metadata file becomes increasingly important because it specifies who can use it (the license) and whom to contact (you!) if there are any problems.

Every package must have a `DESCRIPTION`. In fact, it's the defining feature of a package (RStudio and devtools consider any directory containing `DESCRIPTION` to be a package). To get you started, `devtools::create("mypackage")` automatically adds a bare-bones description file. This will allow you to start writing package without having to worry about the metadata until you need to. The minimal description will vary a bit depending on your settings, but should look something like this:

    Package: mypackage
    Title: What The Package Does (one line, title case required)
    Version: 0.1
    Authors@R: "First Last <first.last@example.com> [aut, cre]"
    Description: What the package does (one paragraph)
    Depends: R (>= 3.1.0)
    License: What license is it under?
    LazyData: true

(If you're writing a lot of packages, you can set global options via `devtools.desc.author`, `devtools.desc.license`, `devtools.desc.suggests`, and `devtools.desc`. See `package?devtools` for more details.)

`DESCRIPTION` uses a simple file format called DCF, the Debian control format. You can see most of the structure in the simple example below. Each line consists of a __field__ name and a value, separated by a colon. When values span multiple lines, they need to be indented:

    Description: The description of a package is usually long,
        spanning multiple lines. The second and subsequent lines
        should be indented, usually with four spaces.


The minimal `DESCRIPTION` file only provides the bare necessities. It doesn't include the two most useful fields: `Imports` and `Suggests`. `Imports` tells R what packages you need. `Suggests` tells R what packages you want to be available. The other fields are described in the remainder of the chapter, and include:

* `Package`: what your package is called.

* `Title` and `Description`: what your package does.

* `License`: who's allowed to use and distribute it

* `Authors@R`: who wrote it

## What does the package need? {#dependencies}

It's the job of the `DESCRIPTION` to list the packages that your package needs to work. R has a rich set of ways of describing potential dependencies. For example, the following lines indicate that your package needs both ggvis and dplyr to work:

    Imports:
        dplyr,
        ggvis

Whereas, the lines below indicate that while your package can take advantage of ggvis and dplyr, they're not required to make it work:

    Suggests:
        dplyr,
        ggvis,

Both `Imports` and `Suggests` take a comma separated list of package names. I recommend putting one package on each line, and keeping them in alphabetical order. That makes it easy to skim.

`Imports` and `Suggests` differ in their strength of dependency:

*   `Imports`: packages listed here _must_ be present for your package to 
    work. In fact, any time your package is installed, those packages will, if 
    not already present, be installed on your computer (`devtools::load_all()`
    also checks that the packages are installed).
    
    Adding a package dependency here ensures that it'll be installed. However, 
    it does _not_ mean that it will be attached along with your package (i.e., 
    `library(x)`). The best practice is to explicitly refer to external 
    functions using the syntax `package::function()`. This makes it very easy to 
    identify which functions live outside your package. This is especially 
    useful when you read your code in the future.
  
    If you use a lot functions from other packages this is rather
    verbose. There's also a minor performance penalty associated with `::` 
    (on the order of 5µs, so it will only matter if you call the function
    millions of times). You'll learn about alternatives in 
    [namespace imports](#imports).

*   `Suggests`: your package can take advantage of packages listed here. Such 
    advantages include specific datasets, the needs of one function in your 
    package, or their use in tests or in building vignettes. 
  
    Packages listed in `Suggests` are not automatically installed along with
    your package. This means that you need to check if the package is available 
    before using it (use `requireNamespace(x, quietly = TRUE)`). There are
    two basic scenarios:
    
    ```{r}
    # You need the suggested package for this function    
    my_fun <- function(a, b) {
      if (!requireNamespace("pkg", quietly = TRUE)) {
        stop("Pkg needed for this function to work. Please install it.",
          call. = FALSE)
      }
    }
    
    # There's a fallback method if the package isn't available
    my_fun <- function(a, b) {
      if (requireNamespace("pkg", quietly = TRUE)) {
        pkg::f()
      } else {
        g()
      }
    }
    ```

When developing packages locally, you never need to use `Suggests`. When releasing your package, using `Suggests` is a courtesy to your users. It frees them from downloading rarely needed packages, and lets them get started with your package as quickly as possible.

The easiest way to add `Imports` and `Suggests` to your package is to use `devtools::use_package()`. This automatically puts them in the right place in your `DESCRIPTION`, and reminds you how to use them.

```{r, eval = FALSE}
devtools::use_package("dplyr") # Defaults to imports
#> Adding dplyr to Imports
#> Refer to functions with dplyr::fun()
devtools::use_package("dplyr", "Suggests")
#> Adding dplyr to Suggests
#> Use requireNamespace("dplyr", quietly = TRUE) to test if package is 
#>  installed, then use dplyr::fun() to refer to functions.
```

### Versioning

If you need a specific version of a package, specify it in parentheses after the package name:

    Imports:
        ggvis (>= 0.2),
        dplyr (>= 0.3.0.1)
    Suggests:
        MASS (>= 7.3.0)

You almost always want to specify a minimum version rather than an exact version (`MASS (= 7.3.0)`). Since R can't have multiple versions of the same package loaded at the same time, specifying an exact dependency dramatically increases the chance of conflicting versions.

Versioning is most important when you release your package. Usually people don't have exactly the same versions of packages installed that you do. If someone has an older package that doesn't have a function your package needs, they'll get an unhelpful error message. However, if you supply the version number, they'll get a error message that tells them exactly what the problem is: an out of date package. 

Generally, it's always better to specify the version and to be conservative about which version to require. Unless you know otherwise, always require a version greater than or equal to the version you're currently using.

### Other dependencies

The are three other fields that allow you to express more specialised dependencies:

* `Depends`: packages listed here require a specific version of R.
  For example, `Depends: R (>= 3.0.1)`. As with packages, it's a good idea to 
  play it safe and to set the version of R to then one you're currently using. 
  `devtools::create()` will do this for you.
  
  Prior to the rollout of namespaces in R 2.14.0, `Depends` was the only way to 
  "depend" on another package. Now, despite the name, you should almost always 
  use `Imports`, not `Depends`. You'll learn why, and when you should still
  use `Depends`, in [namespaces](#namespace).

  In R 3.1.1 and earlier you'll also need to use `Depends: methods` if 
  you use S4. This bug is fixed in R 3.2.0, so methods can go back to
  `Imports`.
  
* `LinkingTo`: packages listed here need to link to or compile against 
  the C code included in another package. You'll learn more about `LinkingTo`
  in [compiled code](#src).
  
* `Enhances`: packages listed here are "enhanced" by your package. Typically, 
  this means you provide methods for classes defined in another packags (a 
  sort of reverse `Suggests`). But it's hard to define what that means, so I 
  don't recommend using `Enhances`.
  
You can also list things that your package needs outside of R in the in `SystemRequirements` field. But this is just a plain text field and is not automatically checked. Think of it as a quick reference; you'll also need to include detailed system requirements (and how to install them) in your README.

### Exercises

* What are the dependencies of ggplot2?

* What does `devtools::revdep()` do? Why might you use it?

## Naming your package {#naming}

The `Package` field lists the name of the package. It should be same as the directory name (and the RStudio project file). For me, the hardest thing about creating a new package is often coming up with a good name. 

### Requirements for a name

There's only one formal requirement: the name can only consist of letters, numbers and `.` (and it must start with a letter and cannot end with a period). Unfortunately this means you can't use `-` or `_` in your package name. I recommend against using `.` in package names because it has confusing connotations (i.e., file extension or S3 method).

### Strategies for creating a name

If you're planning on releasing your package, I think it's worth spending a few minutes to come up with a good name. Here are some recommendations for how to go about it:

* Pick a unique name so you can easily Google it. This makes it easy for 
  potential users to find your package (and associated resources), and it makes 
  it easier for you to see who's using it.

* Avoid using both upper and lower case letters: they make the package name
  hard to type and hard to remember. For example, I can never remember if it's
  `Rgtk2` or `RGTK2` or `RGtk2`.

* Find a word that evokes the problem and modify it so that it's unique:
  `plyr` (generalisation of apply tools), `lubridate` (makes dates and times
  easier), `mutatr` (mutable objects), `classifly` (high-dimensional views of
  classification).

* Use abbreviations: `lvplot` (letter value plots), `meifly` (models explored
  interactively).

* Add an extra R: `stringr` (string processing), `tourr` (grand tours), `httr`
  (HTTP requests).

Some good examples:

* [knitr](http://yihui.name/knitr/): "the package name knitr was coined with 
  weave in mind, and it also aims to be neater."
  
* [analogsea](https://github.com/sckott/analogsea), an R package that talks to
  the digital ocean API.

## What does the package do? {#pkg-description}

The title and description fields describe what the package does. They differ only in length:

* `Title` is a one line description of the package, and is often shown in 
  package listing. It should be plain text (no markup), capitalised like a 
  title, and NOT end in a period. Keep it short: listings will often 
  truncate the title to 65 characters.

* `Description` is more detailed that the title. You can use multiple sentences 
  but you are limited to one paragraph. If your description spans multiple 
  lines (and it should!), each line must be no more than 80 characters wide. 
  Indent subsequent lines with 4 spaces.

The `Title` and `Description` for ggplot2 are:

```
Title: An implementation of the Grammar of Graphics
Description: An implementation of the grammar of graphics in R. It combines 
    the advantages of both base and lattice graphics: conditioning and shared 
    axes are handled automatically, and you can still build up a plot step 
    by step from multiple data sources. It also implements a sophisticated 
    multidimensional conditioning system and a consistent interface to map
    data to aesthetic attributes. See the ggplot2 website for more information, 
    documentation and examples.
```

A good title and description are important, especially if you plan to release your package to CRAN because they appear on the CRAN download page as follows:

```{r, echo = FALSE}
bookdown::embed_png("diagrams/cran-package.png")
```

Because `Description` only gives you a small amount of space to describe what your package does, I also recommend including a `README.md` file that goes into much more depth and shows a few examples. You'll learn about that [README.md](#readme).

### Exercises

* Read the title and description of the packages that you use most frequently.
  What works well? What could be done better?

## Who wrote the package? {#author}

To identify the package's author, and whom to contact if something goes wrong, use the `Authors@R` field. This field is unusual because it contains executable R code rather than plain text. Here's an example:

    Authors@R: person("Hadley", "Wickham", email = "hadley@rstudio.com",
      role = c("aut", "cre"))

```{r}
person("Hadley", "Wickham", email = "hadley@rstudio.com", 
  role = c("aut", "cre"))
```

This command says that both the author (aut) and the maintainer (cre) is Hadley Wickham, and that his email address is `hadley@rstudio.com`. The `person()` function has four main arguments:

* The name, specified by the first two arguments, `given` and `family` (these
  are normally supplied by position, not name). In English cultures, `given` 
  (first name) comes before `family` (last name). In many cultures, this 
  convention does not hold.
  
* The `email` address.

* A three letter code specifying the `role`. There are four important roles:

    * `cre`: the creator or maintainer, the person you should bother 
      if you have problems.
      
    * `aut`: authors, those who have made significant contributions to the 
      package.
    
    * `ctb`: contributors, those who have made smaller contributions, like 
      patches.
    
    * `cph`: copyright holder. This is used if the copyright is held by someone 
      other than the author, typically a company (i.e. the author's employer).
    
    (The [full list of roles](http://www.loc.gov/marc/relators/relaterm.html) is
    extremely comprehensive. Should your package have a woodcutter ("wdc"), 
    lyricist ("lyr") or costume designer ("cst"), rest comfortably that you can 
    correctly describe their role in creating your package.)

If you need to add additional clarification, you can also use the `comment` argument and supply the desired information in plain text.

You can list multiple authors with `c()`:

    Authors@R: c(
        person("Hadley", "Wickham", email = "hadley@rstudio.com", role = "cre"),
        person("Winston", "Chang", email = "winston@rstudio.com", role = "aut"))

Alternatively, you can do this concisely by using `as.person()`:

    Authors@R: as.person(c(
        "Hadley Wickham <hadley@rstudio.com> [aut, cre]", 
        "Winston Chang <winston@rstudio.com> [aut]"
      ))

(This only works well for names with only one first and last name.)

Every package must have at least one author (aut) and one maintainer (cre) (they might be the same person). The creator must have an email address. These fields are used to generate the basic citation for the package (e.g. `citation("pkgname")`). Only people listed as authors will be included in the auto-generated citation. There are a few extra details if you're including code that other people have written. Since this typically occurs when you're wrapping a C library, it's discussed in [compiled code](#src).

As well as your email address, it's also a good idea to list other resources availble for help. You can list URLs in `URL` Multiple URLs are separated with a comma. `BugReports` is the URL where bug reports should be submitted. For example, knitr has:

    URL: http://yihui.name/knitr/
    BugReports: https://github.com/yihui/knitr/issues

You can also use separate `Maintainer` and `Author` fields. I prefer not to use these fields because `Authors@R` offers much richer metadata. 

### On CRAN

The most important thing to note is that your email address (i.e., the address of `cre`) is the address that CRAN will use to contact you about your package. So make sure you use an email address that's likely to be around for a while. Also, because this address will be used for automated mailings, CRAN policies require that this be for a single person (not a mailing list) and that it does not require any confirmation or use any filtering.

## Who can use it? {#license}

The `License` field can be either a standard abbreviation for an open source license, like `GPL-2` or `BSD`, or a pointer to a file contain more information `file LICENSE`. The license is really only important if you're planning on releasing your package. If you don't, you can ignore this section. If you want to make it clear that your package is not open source, use `License: file LICENSE` and then create a file called `LICENSE`, containing (e.g.):

    Proprietary 

    Do not distribute outside of Widgets Incorporated.

Open source software licensing is a rich and complex field. Fortunately, in my opinion, there are only three licenses that you should consider for your R package:

  * [MIT](https://tldrlegal.com/license/mit-license) 
    (v. similar: to BSD 2 and 3 clause licenses). This is a simple and
    permissive license. It lets people use  and freely distribute your code
    subject to only one restriction: the license must always be distributed
    with the code.
  
    The MIT license is a "template", so if you use it, you need 
    `License: MIT + file LICENSE`, and `LICENSE` file that looks like this:
    
    ```
    YEAR: <Year or years when changes have been made>
    COPYRIGHT HOLDER: <Name of the copyright holder>
    ```

  * [GPL-2](https://tldrlegal.com/license/gnu-general-public-license-v2) or 
    [GPL-3](https://tldrlegal.com/license/gnu-general-public-license-v3-(gpl-3)). 
    These are "copy-left" licenses. This means that any one who distributes your
    code in a bundle must license the whole bundle in a GPL-compatible way. 
    Additionally, anyone who distributes modified versions of your code 
    (derivative works) must also make the source code available. GPL-3 is a 
    little stricter than GPL-2, closing some older loopholes.

  * [CC0](https://tldrlegal.com/license/creative-commons-cc0-1.0-universal). 
    It relinquishes all your rights on the code and data so that it can be 
    freely used by any one for any purpose. This is sometimes called putting it 
    in the public domain, a term which is neither well-defined nor meaningful in 
    all countries.
  
    This license is most appropriate for data packages. Data, at least in the US, 
    is not copyrightable, so you're not really giving up much. This 
    license just makes this point clear.
  
If you'd like to learn more about other common licenses, Github's [choosealicense.com](http://choosealicense.com/licenses/) is a good place to start. Another good resource is <https://tldrlegal.com/>, which explains the most important parts of each license. If you use a license other than the three I suggest, make sure you consulte the "Writing R Extensions" section on [licensing][R-exts].

If your package includes code that you didn't write, you need to make sure you're in compliance with its license. Since this occurs most commonly when you're including C source code, it's discussed in more detail in [compiled code](#src).

### On CRAN

If you want to release your package to CRAN, you must pick a standard license. Otherwise it's difficult for CRAN to determine whether or not it's legal to distribute your package! You can find a complete list of licenses that CRAN considers valid at <https://svn.r-project.org/R/trunk/share/licenses/license.db>.

```{r, results='asis', echo = FALSE, eval = FALSE}
licenses <- read.dcf(file.path(R.home("share"), "licenses", "license.db"))
licenses <- as.data.frame(licenses, stringsAsFactors = FALSE)
licenses <- licenses[order(licenses$Name, licenses$Version), ]
licenses[is.na(licenses)] <- ""
has_abbrev <- subset(licenses, Abbrev != "")
knitr::kable(has_abbrev[c("Name", "Version", "Abbrev")], row.names = FALSE)
```

## Version {#version}

An R package version is a sequence of at least two integers separated by either `.` or `-`. For example, `1.0` and `0.9.1-10` are valid versions, but `1` or `1.0-devel` are not. You can parse a version number with `numeric_version`.

```{r}
numeric_version("1.9") == numeric_version("1.9.0")
numeric_version("1.9.0") < numeric_version("1.10.0")
```

For example, a package might have a version 1.9. This version number is considered by R to be the same as 1.9.0, less than version 1.9.2, and all of these are less than version 1.10 (which is version "one point ten", not "one point one zero). R uses version numbers to determine whether package dependencies are satisfied. A package might, for example, import package `devtools (>= 1.9.2)`, in which case version 1.9 or 1.9.0 wouldn't work.

The version number of your package increases with subsequent releases of a package, but it's more than just an incrementing counter -- the way the number changes with each release can convey information about what kind of changes are in the package.

I don't recommend taking full advantage of R's flexiblity. Instead always use `.` to separate version numbers. 

* A released version number consists of three numbers, `<major>.<minor>.<patch>`. 
  For version number 1.9.2, 1 is the major number, 9 is the minor number, and 
  2 is the patch number. Never use versions like `1.0`, insted always spell
  out the three components, `1.0.0.`

* An in-development package has a fourth component: the development version.
  This should start at 9000. For example, the first version of the package
  should be `0.0.0.9000`. There are two reasons for this recommendation:
  first, it makes it easy to see if a package is released or in-development,
  and the use of the fourth place means that you're not limited to what the
  next version will be. `0.0.1`, `0.1.0` and `1.0.0` are all greater than 
  `0.0.0.9000`.
  
    Increment the development version, e.g. from `9000` to `9001` if you've
    added an important feature that another development package needs to depend 
    on.

This advice here is inspired in part by [Semantic Versioning](http://semver.org) and by the [X.Org](http://www.x.org/releases/X11R7.7/doc/xorg-docs/Versions.html) versioning schemes. Read them if you'd like to understand more about the standards of versioning used by many open source projects.hc

We'll come back to version numbers in the context of releasing your package, [picking a version number](#release-version). For now, just remember that the first version of your package should be `0.0.0.9000`.

## Other components {#description-misc}

A number of other fields are described elsewhere in the book:

* `Collate` controls the order in which R files are sourced. This only
  matters if your functions have side-effects; most commonly because you're
  using S4. This is described in more depth in [documenting S4](#man-s4).

* `LazyData` makes it easier to access data in your package. Because it's so 
  important, it's included in the minimal description. It's described in more
  detail in [external data](#data).

There are actually many other rarely, if ever, used fields. A complete list can be found in the "The DESCRIPTION file" section of the [R extensions manual][description]. You can also create your own fields to add additional metadata. The only restrictions are that you shouldn't use existing names and that, if you plan to submit to CRAN, the names you use should be valid English words (so a spell-checking NOTE won't be generated).

[R-exts]: http://cran.r-project.org/doc/manuals/R-exts.html#Licensing
