---
title: Package basics
layout: default
output: bookdown::html_chapter
---

# Package metadata {#description}

The job of the `DESCRIPTION` file is to store important metadata about your package. In the early stages of package development, the most important thing that the `DESCRIPTION` does is record which other packages your package needs. It becomes becomes increasingly important as you get closer to releasing your package to others because it also lays out who can use it (the license), and who to contact (you!) if there are any problems.

## A minimal `DESCRIPTION` file

Every package must have a `DESCRIPTION`, and in fact it's the distinguishing feature of a package - RStudio and devtools consider any directory containing a `DESCRIPTION` to be a package. To get you started, `devtools::create()` automatically adds a minimal description. This means you don't need to worry about the `DESCRIPTION` in the early stages of package development. 

The minimal description will vary a bit depending on some system settings, but will look something like this:

    Package: package-name
    Title: What the package does (short line)
    Version: 0.1
    Authors@R: "First Last <first.last@example.com> [aut, cre]"
    Description: What the package does (paragraph)
    Depends: R (>= 3.1.0)
    License: What license is it under?
    LazyData: true

The `DESCRIPTION` uses a simple file format called DCF, the Debian control format. You can see most of the structure in this simple example. Each line consists of a __field__ name and a value, separated by a colon. When values span multiple lines, they need to be indented:

    Description: The description of a package is usually long,
        spanning multiple lines. The second and subsequent lines
        should be indented, usually with four spaces.

The minimal `DESCRIPTION`, but provides the bare necessities to be valid. This is the critical subset of package metadata: what it's called (`Package`), what it does (`Title`, `Description`), what version it is (`Version`), who's allowed to use and distribute it (`License`), and who wrote it (`Author@R`).

In the rest of the chapter, you'll learn what these fields do. We'll start with package, title and description, since they're fairly next. Next you'll learn about the two most important fields. They are not included in the minimal description, but are `Imports` and `Suggests`: these are the fields that tell R that you need (for imports) or want (for suggests) additional packages to be available. You'll learn about those fields in the next section, and then we'll go on to describe the other important fields in the rest of the chapter.

## What is your package called? {#naming}

The `Package` field defines the name of the package. This should be same as the name of directory. Often the hardest thing about creating a new package is coming up with a good name. There's only one formal requirement: the package name can only consist of letters, numbers and `.`. It must start with a letter and cannot end with a period. Unfortunately this means you can't use `-` or `_` in your package name. I recommend against using `.` to in package names because the other connotations (i.e. file extension or S3 method) are confusing.

If you're planning on releasing your package to the wild, I think it's worth spending a few minutes to come up with a good name. I have two recommendations:

* Pick a unique name so you can easily google it. This makes it easy for 
  potential users to find your package, and it makes it easier for you to
  see who's using your package.

* Avoid using both upper and lower case letters: they make the package name
  hard to type and hard to remember. For example, I can never remember if it's
  `Rgtk2` or `RGTK2` or `RGtk2`.

Some strategies I've used in the past to create packages names:

* Find a name evocative of the problem and modify it so that it's unique:
  `plyr` (generalisation of apply tools), `lubridate` (makes dates and times
  easier), `mutatr` (mutable objects), `classifly` (high-dimensional views of
  classification).

* Use abbreviations: `lvplot` (letter value plots), `meifly` (models explored
  interactively).

* Add an extra R: `stringr` (string processing), `tourr` (grand tours), `httr`
  (HTTP requests).

## What does the package do?

There are two fields that describe the package. They differ in length:

* `Title` is a one line description of the package, and is often shown in 
  package listing. It should be plain text (no markup), be capitalised like a 
  sentence, but not end in a period. Keep it short: listings will often 
  truncate the title to 65 characters.

* `Description` is more detailed: you can use multiple sentences, but still 
  only one paragraph. The most common place to see the description is on the
  CRAN download page.
  
  If your description spans multiple lines (and it should!), keep each
  line at most 80 characters wide, and indent subsequent lines with 4 spaces.

The `Title` and `Description` for ggplot2 are:

```
Title: An implementation of the Grammar of Graphics
Description: An implementation of the grammar of graphics
    in R. It combines the advantages of both base and
    lattice graphics: conditioning and shared axes are
    handled automatically, and you can still build up a
    plot step by step from multiple data sources. It also
    implements a sophisticated multidimensional
    conditioning system and a consistent interface to map
    data to aesthetic attributes. See the ggplot2 website
    for more information, documentation and examples.
```

And these are rendered on the CRAN download page as follows:

```{r, echo = FALSE}
bookdown::embed_png("diagrams/cran-package.png")
```

### Exercises

* Read the title and description of the packages that you use most commonly.
  What do they do well? What could they do better?

## What does the package need?

It's the job of the `DESCRIPTION` to list which other packages that your package needs to work. R has a rich set of ways of describing potential dependencies. The two most important are:

*   `Imports`: your package requires these packages to work. This ensures the
    package is available, but does not automatically load all functions into your
    workspace. Instead, it's best practice to explicitly refer to external 
    functions using the syntax `package::function()`. When reading your code in
    the future, this makes it very easy to identify functions that live outside
    your package.
  
    Adding packages to `Imports` ensures that they are always available. When
    your package is installed, any packages that are listed in Imports but are
    not currently installed will be automatically installed.
  
    If you use functions from another package extensively, this can be a bit
    verbose and add unnecessary mental reading weight. There is also a 
    minor performance penalty (normally only important if you're inside a loop
    that's being called millions of times). You'll learn about alternatives in
    [namespace imports](#imports).

*   `Suggests`: you don't absolutely need these packages but they might be
    helpful. Maybe they provide datasets for examples, or only used by one
    function in your package. 
  
    Packages listed in `Suggests` are not automatically installed when your 
    package is installed. This means that you need to check if the package is
    available before calling any functions in it. Unfortunately base R does 
    not come with a function to check if a package is installed, so you'll
    need to add this simple function to your package:
  
    ```{r}
    is_installed <- function(pkg, version = 0) {
      system.file(package = pkg) != "" && packageVersion(pkg) > version
    }
    ```
    
    Then your code will look like this:
    
    ```{r}
    my_fun <- function(a, b) {
      if (!is_installed("pkg")) {
        stop("Pkg needed for this function to work. Please install it.",
          call. = FALSE)
      }
      x <- pkg::fun1(a, b)
      pkg::fun2(x, 10)
    }
    ```
    
    As we'll learn later, any package that you need for testing or vignettes
    must be listed in (at least) suggests.

The easiest way to add dependencies to your package is to use `devtools::use_package()`. This automatically adds them in the right place in your `DESCRIPTION`, and reminds you how to use them.

```{r, eval = FALSE}
devtools::use_package("Rcpp") # Defaults to imports
devtools::use_package("Rcpp", "Suggests")
```

### Versioning

It's also possible to specify version information about the required packages. You can specify a minimal required version like `MASS (>= 7.3.0)`.  You can specify exact version dependency with `MASS (= 7.3.0)`, but this is not recommended - since R can't have multiple versions of the same package loaded at the same time, specifying an exact dependency dramatically increases the chance of conflicting versions that can't be resolved.

Versioning is particularly important if you want to release your package. Often people will have an older version of a package loaded, and they will get an unhelpful error if you rely on new features without being explicit about the version. Generally, I think it's better to be conservation about version specifications - unless you know otherwise, add the version that you're currently using.

### Other depencies

Three others are useful in more specialised circumstances:

* `Depends`: used to specify version of R needed, and for a few other 
  special cases. Otherwise mostly historical and should not be used.
  This is discussed in more detail in [namespaces](#namespace).

* `LinkingTo`: needs to link to or compile against the C code included in
  this package. This is discussed in more detail in [compiled code](#src).
  
* `Enhances`: these packages are "enhanced" by your package, typically because
  you provide methods for classes defined in the package. It's a sort of reverse
  suggests. But it's hard to define what that actually means, so I recommend 
  that you don't use it.
  
http://cran.r-project.org/doc/manuals/R-exts.html#Package-Dependencies

You can also list other requirements (outside of R) in `SystemRequirements`. But this is basically only useful as a quick reference - you'll also need to include detailed information in a readme.

### Exercises

* What are the dependencies of ggplot2?

* What does `devtools::rev_dep()` do? Why might you use it?

## Who wrote the package?

Instead of `Maintainer` and `Author`, you can `Authors@R`, which takes a vector of `person()` elements.  Each person object specifies the name of the person and their role in creating the package:

* `aut`: full authors who have contributed much to the package

* `ctb`: people who have made smaller contributions, like patches.

* `cre`: the package creator/maintainer, the person you should bother if you
  have problems
  
(The [full list of roles](http://www.loc.gov/marc/relators/relaterm.html) is extremely comprehensive. Should your package have a woodcutter ("wdc"), lyricist ("lyr") or costume designer ("cst"), rest comfortably that you can correctly ascribe their role in your package.)

Other roles are listed in the help for person. Using `Authors@R` is useful when your package gets bigger and you have multiple contributors that you want to acknowledge appropriately. The equivalent `Authors@R` syntax for plyr would be:

      Authors@R: person("Hadley", "Wickham", role = c("aut", "cre"))

As well as providing your email address, it's also a good idea to list other resources avaialble for help. You can list urls in `URL` Multiple urls can be separated with a comma. `BugReports` takes a url to where bug reports should be submitted.

## Who can use it? (`LICENSE`)

`License`: a standard abbreviation for an open source license, like `GPL-2` or `BSD`.

The license is only really important if your planning on releasing your package. If you don't, you can ignore this section.

Open source software licensing is a rich and complex field. There are many many options to choose from. I'm not a lawyer, and if you're really concerned you should consult one. Instead, I'm going to give you some fairly sweeping advice that holds for most cases. In my opinon, there are really only three licenses that you need to worry about for R packages:

* [MIT](https://tldrlegal.com/license/gnu-general-public-license-v2) 
  (v. similar: to BSD 2 and 3 clause licenses): this is simple and
  permissive license. The only restriction is that if you distribute software
  that uses the MIT license, you must also distribute the license.
  
  The MIT license is a "template", so if you use it, you need 
  `License: MIT + file LICENSE`, and `LICENSE` file that looks like this:
  
    ```
    YEAR: 2014
    COPYRIGHT HOLDER: Hadley Wickham, RStudio
    ```

* [GPL-2](https://tldrlegal.com/license/gnu-general-public-license-v2) or 
  [GPL-3](https://tldrlegal.com/license/gnu-general-public-license-v3-(gpl-3)): 
  a "copy-left" licenses. These licenses are stricter
  and require any who distributes your code, or modified versions of you code
  (derivative works), to also make the source code avaialble. GPL-3 is a little
  stricter than GPL-2, attempting to close some loopholes.

* `CC0`: This is most appropriate for data packages. This relinquishes all your
  rights on the code and data so that it can be freely used by any one for any
  purpose (this is sometimes called putting it in the public domain, although
  that term is not well-defined, and not meaningful in all countries). 
  
  Data, at least in the US, is not copyrightable anyway, so you're not
  really giving up much. This license just makes it clear.
  
If you'd like to learn more about others, Github's [http://choosealicense.com/licenses/](choosealicense.com) is a good place to start for the most popular licenses. Another good resource is <https://tldrlegal.com/>, which explains the most important parts of each license.

If you are using a non-standard license, put `file LICENSE` and then include the full text of the license in a `LICENSE`.

If you include code written by other people, you need to make sure you're include their license details. Since this occurs most commonly when you're including C source code, it's discussed in more detail there.

### On CRAN

If you want to release your package to CRAN, you need to pick a standard license. Otherwise it's difficult for CRAN to determine whether or not it's legal for them to distribute your package!

A complete list of valid licenses for cran can be found at <https://svn.r-project.org/R/trunk/share/licenses/license.db>.

```{r, results='asis', echo = FALSE, eval = FALSE}
licenses <- read.dcf(file.path(R.home("share"), "licenses", "license.db"))
licenses <- as.data.frame(licenses, stringsAsFactors = FALSE)
licenses <- licenses[order(licenses$Name, licenses$Version), ]
licenses[is.na(licenses)] <- ""
has_abbrev <- subset(licenses, Abbrev != "")
knitr::kable(has_abbrev[c("Name", "Version", "Abbrev")], row.names = FALSE)
```


## Other components

* `Version`: the version number, which should be of the the form 
  `major.minor.patchlevel`. This is most important when releasing your package. 
  See [version numbers](#version) for more deatils.

* `LazyData`. Described in [external data](#data).

There are a number of other less commonly used fields like `KeepSource`, `OS_type` and `Language`. A complete list can be found in the [R extensions manual][description].

## Collation order {#collate}

R loads files in alphabetical order. Unfortunately not every alphabet puts letters in the same order, so you can't rely on alphabetic ordering if you need one file loaded before another. The order in which files are loaded doesn't matter for most packages. But if you're using S4, you'll need to make sure that classes are loaded before subclasses and generics are defined before methods.

Rather than relying on alphabetic ordering, roxygen2 provides an explicit way of saying that one file must be loaded before another: `@include`. The `@include` tag gives a space separated list of file names that should be loaded before the current file:

```{r, eval = FALSE}
#' @include class-a.r
setClass("B", contains = "A")
```

If any `@include` tags are present in the package, roxygen2 will set the `Collate` field in the `DESCRIPTION`, which ensures that files are always loaded in the same order.

[description]: http://cran.r-project.org/doc/manuals/R-exts.html#The-DESCRIPTION-file
