---
title: Git and github
layout: default
output: bookdown::html_chapter
---

# Git and github {#git}

  * Do you have a directory full of files like `my-script.R`, `my-script-1.R`,
    `myscript-2-I-really-hope-this-works.R`, `myscript-FINALLY.R` and so on?
    Git takes care of managing multiple versions of your code so that you
    can easily see what's change, and revert any mistakes that you've made.

  * It makes it easy for other people to install your package. Any R user can 
    get your package with at most two lines of code:
    
    ```{r, eval = FALSE}
    install.packages("devtools")
    devtools::install_github("username/packagename")
    ```
    
  * Not only can other people use your code, they can also suggest improvements 
    with pull requests. Once you've experienced your first pull request, you'll
    never want to go back to developing code any other way.
    
  * Github makes it possible to collaboratively code a package with other 
    people. As long as you're working on different parts of the file,
    git can figure out how to combine your changes. If you are working on
    the same part of the file, git provides tools to help you choose
    between the conflicting changes.

  * Track issues.

At first, working in the open seems a little scary. Do you really want the whole world seeing your crappy code? Don't worry, nobody is judging you - everybody rights bad code at some point in their lives, and most people are more interested in helping you write better code than making fun of your existing efforts. Once you get used to it, you'll find that coding in the open is tremendously empowering as it helps other people help you.

You can do many of the same things with other tools (like svn or bazaar or mercurial) and other websites (like gitlab and bitbucket). But I think git and github is the friendliest system for new developers, not least because it's the most popular, which means every possible problem has already been asked and answered on StackOverflow. Indeed, StackOverflow is a vital part of git - when you have a problem that you don't know how to solve, SO should be your first recourse.

This chapter describes git and github together, making no attempt to separate the features of each. The goal is to give you the absolute minimum you need to know to use git for an R package. After reading this chapter and working with git for a while, it's highly likely that you'll want to learn more. Some good resource are:

* Github help, <https://help.github.com>. This not only teaches you about
  github, but also many of the features of git as well.
  
* If you'd like to practice using git from the command line, try 
  <https://try.github.io>. It steps you through a number of challenges in
  a virtual shell.

* If you'd like to learn more about the details of git, read 
  [Pro Git](http://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub.

RStudio provides many tools to make your day-to-day use of git as easy as possible. However, there are a huge number of git commands, and they're not at all available in the IDE. That means you'll need to run a handful of commands from a shell (aka a console), especially when you're setting up, dealing with merge conflicts and getting out of jams. The easiest way to get to a shell is Tools > Shell.

Git and backups. Git is not a replacement for a backup system.

## Initial set up

If you've never used git or github before, you'll need to do a little initial setup:

1.  Install git:

    * Windows: <http://msysgit.github.io/>
    * OS X: <http://code.google.com/p/git-osx-installer/>
    * Debian/Ubuntu: `sudo apt-get install git-core`

1.  Sign up for github: <https://github.com>.

1.  In a shell, run:

    ```bash
    git config --global user.name "<YOUR NAME>"
    git config --global user.email "<YOUR EMAIL ADDRESS>"
    ```
    
1.  If needed, generate ssh keys. You can check if you have an ssh key
    already by running:
    
    ```{r, eval = FALSE}
    file.exists("~/.ssh/id_rsa.pub")
    ```

    If it's `FALSE`, you'll need to create one. You can follow the [github 
    instructions](https://help.github.com/articles/generating-ssh-keys) or
    use RStudio. Go to RStudio preferences, choose the Git/SVN panel, then 
    click "Create RSA key...":
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-config.png", dpi = 220)
    ```

## Create a local git repository

Now that you have git installed and configured, you need to initialise a local git repository for your package. This repository (or repo for short) exists only on your computer. 

There are two ways to do this:

  * In RStudio, go to project options, then the Git/SVN panel. Change the 
    "Version control system" from "None" to "Git":
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-proj-config.png", dpi = 220)
    ```
    
    You'll then be prompted to restart RStudio.

  * In a shell, run `git init`. Restart RStudio and reopen your package.

Once git has been activated in a project, you'll see two new components in the IDE:

  * The git pane. This appears in the top-right by default. It shows you what
    files have changed since you last committed, and exposes the most important
    git commands as buttons.
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-pane.png", dpi = 220)
    ```
    
  * The git dropdown in the toolbar. This exposes git and github commands
    useful for working with the current file:
  
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-dropdown.png", dpi = 220)
    ```

## See what's changed

The RStudio git pane shows you at a glance what you've changed since the last time you committed ("saved") your changes. This is equivalent to running `git status` in the shell. Each modified file is listed, along with an icon summarising the change:

* Unknown: `r bookdown::embed_png("screenshots/git-unknown.png", dpi = 220)`.
  You'll see this when you first add a file to git
  when you first add a new file, git doesn't 

* Deleted: `r bookdown::embed_png("screenshots/git-deleted.png", dpi = 220)`

* Modified: `r bookdown::embed_png("screenshots/git-modified.png", dpi = 220)`.
  You can get more details about how the files have been modified with a 
  "diff", `r bookdown::embed_png("screenshots/git-diff.png", dpi = 220)`.
  This opens a new window showing the detailed __diff__erences: removed text
  is shown in red and new text is shown in green. (This is equivalent to running
  `git diff` in the shell).

## Add and commit files

The fundamenetal unit of work in git is a __commit__. A commit is a snapshot of the state of your code at a fixed point in time. A commit is a checkpoint in the history of your code - it's fixed and unchanging, and you can go back to it in time.  Each commit is given a unique id, called a sha. 

Commit is used in two senses: 

* A noun: a commit is a snapshot in time.
* A verb: commit you code to create a new snapshot.

I'll try to avoid confusing sentences like you commit code to create a commit.

There are two steps to creating a new commit:

1.  You __stage__ files, telling git that you want to include them in the
    next commit. On the command line, this is done with `git add` for new files
    and `git rm` for old files.
  
1.  You __commit__ the staged files, describe the changes with a message.
    This is equivalent to `git commit` on the command line.

You perform both these steps in the same place in RStudio: the commit window. Open the commit window by clicking `r bookdown::embed_png("screenshots/git-commit.png", dpi = 220)`. This opens three-pane window. This one shows the status while I was working on this chapter.

```{r, echo = FALSE}
bookdown::embed_png("screenshots/git-commit-window.png", dpi = 220)
```

* The top-left pane is the status, the same as the git pane.

* The top-right pane is the commit message, where you provide a human readable
  message describing the changes you've made in this commit. More on that 
  shortly.

* The bottom pane shows the diff for the currently selected file. 

To make a new commit:

1.  Save your current file.

1.  Click commit or press `Ctrl + Cmd + M`.

1.  Stage the files that you want to be included. If you want to include all 
    the files, press Cmd + A, then click 
    `r bookdown::embed_png("screenshots/git-stage.png", dpi = 220)`.
    As you stage each file, you'll notice that its status will change. There 
    are two columns: staged (on the left) and unstaged (on the right). You'll 
    see two new statuses:

      * Added: `r bookdown::embed_png("screenshots/git-added.png", dpi = 220)`: once
        you've staged a new file, git now knows that you've added it to the repo.
      
      * Renamed: `r bookdown::embed_png("screenshots/git-renamed.png", dpi = 220)`:
        If you rename a file, git initially sees it as a delete and an add. If you
        stage the new and old files, git will recognise it as a rename.

1.  Write a commit message (top-right panel). The first line of a commit message 
    should be 50 characters or less and summarise the changes you've made. For
    complicate commits, you can follow it with a blank line and then a paragraph 
    or bulleted list providing more detail. Write messages in imperative, like
    you're telling someone what to do: "fix this bug", not "fixed this bug".
    
    The purpose of the commit message is when other people (including future 
    you!) are skimming the commits it describes what's changed. It's worthwhile 
    to spend a little effort on your commit messages, so that if you do need to 
    look at history, you can easily see. But don't spend too much time!  
    Most important when you're on a team.

1.  Click Commit.

Often there are files that you don't want to include in the repository. They might be transient artefacts (like stuff you get when building latex files, or compiling), or they might be too big, or they might be generated on demand. Instead of not-staging them each time, you should add them to `.gitignore` to prevent them from every being added.  The easiest way to do this is to right-click on the file in the git pane and select `Ignore`. If you want to ignore multiple files, you can use a wildcard like `*.png`.

## Undo a mistake

* If you made a mistake in a commit, you can modify the previous commit by 
  clicking `r bookdown::embed_png("screenshots/git-commit-amend.png", dpi = 220)`. 
  There are more advanced history modifying features available from the command
  line. See [rewriting history](http://git-scm.com/book/en/v2/Git-Tools-Rewriting-History)
  for details.
  
* If you want undo the changes you've made to a file, right click on it in 
  the git pane and select "revert". BEWARE: this will undo all local changes
  and you can't recover them.

## Connect your repo to github

Most of the time you work locally. This is really convenient because you don't need internet access to use git - you can keep on committing changes even when you're on a plane, and then later synchronise with thyour 

To connect github to an existing package:

1.  Create a new repo: <https://github.com/new>

1.  ```
    git remote add origin git@github.com:rstudio/ggcomp.git
    git push -u origin master
    ```

(`use_github()` will be coming soon to a devtools near you)

Once you've put your package on github, make sure to update your `DESCRIPTION` to include links to your repo, e.g.:

```
URL: http://github.com/hadley/dplyr
BugReports: http://github.com/hadley/dplyr/issues
```

## Synchronise with github

Use push to send your changes to github. Use pull to get them back.

How does Rstudio deal with conflicts?

* `git push`, `git pull`

## History

* History view
* View on github
* Blame on github

## Issues

I recommend that you use them. Don't go too crazy with tags and milestones.

Close issues with commit messages: `Close #n`, `@xyz`. etc.

Good idea to add a bullet point to `NEWS.md`. Describe more in user terms. Commit message describes in developer terms.

## Pull requests

### Handling

### Making

* Branch
