---
title: Git and github
layout: default
output: bookdown::html_chapter
---

# Git and github {#git}

If you're serious about software development, the most important tool to learn is git. Git is a software tool revision management tool which allows you to track changes to your code and share those changes with others. Git is most useful when combined with [github](http://github.com), a website that allows you to share your code with the world, solicit code improvements (pull requests) and track issues.

You can do many of the same things with other tools (like svn or bazaar or mercurial) and other websites (like gitlab and bitbucket). But I think git and github is the friendliest system for new developers, not least because it's the most popular, which means every possible problem has already been asked and answered on StackOverflow. 

Why use git and github?

  * Do you have a directory full of files like `my-script.R`, `my-script-1.R`,
    `myscript-2-I-really-hope-this-works.R`, `myscript-FINALLY.R` and so on?
    Git takes care of managing multiple versions of your code so that you
    can easily see what's change, and revert any mistakes that you've made.
    (But note that git isn't a substitute for a backup sytem, you should be
    use git with your backups, not instead of).

  * It makes it easy for other people to install your package. Any R user can 
    get your package with at most two lines of code:
    
    ```{r, eval = FALSE}
    install.packages("devtools")
    devtools::install_github("username/packagename")
    ```
    
    (The first line isn't necesary if they already have devtools installed.)
    
  * Not only can other people use your code, they can also suggest improvements 
    with pull requests, patches to your code that fixb bugs or implement new
    functionality. Once you've experienced your first pull request, you'll
    never want to go back to developing code any other way.
    
  * Github makes it possible to collaboratively code a package with other 
    people. As long as you're working on different parts of the file,
    git can figure out how to combine your changes. If you are working on
    the same part of the file, git provides tools to help you choose
    between the conflicting changes.

  * Track issues (bugs and feature requests) in one central location. You
    can discuss the problem and propose solutions. When you resolve an 
    issue you can connect it to the exact change that solved the problem,
    which is very useful when you come back to it in the future.

  * The Github site for your repo makes a great minimal website for your 
    package. Readers can easily browse code and markdown files are rendered
    as html. This makes the source code for your package much more accessible
    than downloading the `tar.gz` from github.

  * Github releases

At first, working in the open seems a little scary. Do you really want the whole world seeing your crappy code? Don't worry, nobody is judging you - everybody rights bad code at some point in their lives, and most people are more interested in helping you write better code than making fun of your existing efforts. Once you get used to it, you'll find that coding in the open is tremendously empowering as it helps other people help you.

This chapter describes git and github together, making no attempt to separate the features of each. The goal is to give you the absolute minimum you need to know to use git for an R package. After reading this chapter and working with git for a while, it's highly likely that you'll want to learn more. Some good resource are:
  
* If you'd like to practice using git from the command line, try 
  <https://try.github.io>. It steps you through a number of challenges in
  a virtual shell.

* Github help, <https://help.github.com>, not only teaches you about
  github, but also has good tutorials on many git features.

* If you'd like to learn more about the details of git, read 
  [Pro Git](http://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub.

What you won't learn here:

* any history modifying changes like (`git rebase` or `git pull --rebase`).

StackOverflow is a vital part of git - when you have a problem that you don't know how to solve, SO should be your first resource. It's highly likely that some one has had exactly the same problem as you, and there will be a variety of approaches to choose from.

RStudio provides many tools to make your day-to-day use of git as easy as possible. However, there are a huge number of git commands, and they're not at all available in the IDE. That means you'll need to run a handful of commands from a shell (aka a console), especially when you're setting up, dealing with merge conflicts and getting out of jams. The easiest way to get to a shell is Tools > Shell. It's also important to be familiar with using git from the command line because if you're searching for problems, you'll need to know what the standard commands are.

## Initial set up {#git-init}

If you've never used git or github before, you'll need to do a little initial setup:

1.  Install git:

    * Windows: <http://msysgit.github.io/>.
    * OS X: <http://code.google.com/p/git-osx-installer/>.
    * Debian/Ubuntu: `sudo apt-get install git-core`.

1.  Tell git your name and email address. These are used to label each commit
    so that when you start collaborating with others, it's clear who made 
    each change. In a shell, run:

    ```bash
    git config --global user.name "<YOUR NAME>"
    git config --global user.email "<YOUR EMAIL ADDRESS>"
    ```

1.  Create a free account on github: <https://github.com>. Use the same
    name and email address as the previous step.
    
1.  If needed, generate ssh keys. You can check if you have an ssh key
    already by running:
    
    ```{r, eval = FALSE}
    file.exists("~/.ssh/id_rsa.pub")
    ```

    If it's `FALSE`, you'll need to create one. You can follow the [github 
    instructions](https://help.github.com/articles/generating-ssh-keys) or
    use RStudio. Go to RStudio preferences, choose the Git/SVN panel, then 
    click "Create RSA key...":
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-config.png", dpi = 220)
    ```
    
1.  Add your public key to <https://github.com/settings/ssh>. The
    easiest way to get the key is to click "View public key" in the
    Git/SVN preferences pane (as shown above).

## Create a local git repository {#git-init}

Now that you have git installed and configured, you need to initialise a local git repository for your package. This repository (repo for short) exists only on your computer. This is different to old systems like svn which were always connected to a central server. Git makes a clear distinction between local and remote repos; you mostly work locally then push your changes to a remote repo when you're ready to share your work with others.

There are two ways to create a new git repo:

  * In RStudio, go to project options, then the Git/SVN panel. Change the 
    "Version control system" from "None" to "Git":
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-proj-config.png", dpi = 220)
    ```
    
    You'll then be prompted to restart RStudio.

  * In a shell, run `git init`. Restart RStudio and reopen your package.

Once git has been activated in a project, you'll see two new components in the IDE:

  * The git pane, which is shown the top-right (by default). It shows you what
    files have changed since you last committed, and exposes the most important
    git commands as buttons.
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-pane.png", dpi = 220)
    ```
    
  * The git dropdown in the toolbar. This exposes git and github commands
    useful for working with the current file:
  
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-dropdown.png", dpi = 220)
    ```

With the repository set up, you can now:

* See what you've changed: [git status](#git-status).
* Add and commit files: [git commit](#git-commit).
* Undo mistakes: [undo](#git-undo)

To get more out of git, you'll need to connect your repo with github:

* [initialising github](#github-init)
* working with others, [`git pull`](#git-pull)
* review history, [`git log`](#git-log)

## See what's changed {#git-status}

The RStudio git pane shows, at a glance, what's changed. Each added, modified or deleted file is listed, along with an icon summarising the change:

* `r bookdown::embed_png("screenshots/git-modified.png", dpi = 220)`, 
  __Modified__. You've changed the contents of the file.

* `r bookdown::embed_png("screenshots/git-unknown.png", dpi = 220)`, 
  __Untracked__. You've added a new file that git hasn't seen before.

* `r bookdown::embed_png("screenshots/git-deleted.png", dpi = 220)`, 
  __Deleted__. You've deleted a file.
  
You can get the same information (in a slightly different format) in the shell by running `git status`. 

For text files, you can get more details about modifications a "diff", `r bookdown::embed_png("screenshots/git-diff.png", dpi = 220)`. This opens a new window showing the detailed **diff**erences:

```{r, echo = FALSE}
bookdown::embed_png("screenshots/git-diff-window.png")
```

Red blocks show removed text and green blocks show added text. Context, lines surrounding the change, are shown in grey, and help you to place the change in the context of complete file. You can get a similar display in the shell by running `git diff`. 

## Add and commit files {#git-commit}

The fundamental unit of work in git is a __commit__. A commit is a snapshot of the state of your code at a fixed point in time. You can think of a commit as a set of changes: what files did you add, edit and delete? Committing code is the most common git operation. You'll create many commits each day. A commit is made up of:

* A unique identifier called a sha.

* A human-readable commit message.

* A parent, the change that came before this one. (Commits can have multiple
  parents, as you'll learn in XYZ)

* A changeset describing what files were added, modified and changed.

Commit can be a confusing word because it's used in two senses: 

* As a noun: a commit is a snapshot of your code.
* As a verb: commit your changes to create a new snapshot.

I'll try to avoid confusing sentences like you commit code to create a commit.

Creating a commmit occurs in two stages:

1.  You __stage__ files, telling git that you want to include them in the
    next commit. In the shell you use `git add` for new and modified files,
    and `git rm` for deleted files. 
  
1.  You __commit__ the staged files, describe the changes with a message.
    In the shell you use `git commit`.

You perform both these steps in the same place in RStudio: the commit window. Open the commit window by clicking `r bookdown::embed_png("screenshots/git-commit.png", dpi = 220)`, or by pressing Ctrl + Alt + m. Here's a screenshot I took while writing this chapter:

```{r, echo = FALSE}
bookdown::embed_png("screenshots/git-commit-window.png", dpi = 220)
```

There are three panes:

* The top-left pane shows the current status, the same as the git pane in the 
  main RStudio window.

* The bottom pane shows the changes made (the "diff") in the file currently 
  selected in the status pane.

* The top-right pane is where you'll enter the commit message, a human 
  readable message summarising the changes made in the commit. More on 
  that shortly.

(Yes, this is exactly the same window you see when clicking `r bookdown::embed_png("screenshots/git-diff.png", dpi = 220)`!)

To make a new commit:

1.  __Save your changes__.

1.  __Open the commit window__ by clicking 
    `r bookdown::embed_png("screenshots/git-commit.png", dpi = 220)` or 
    pressing `Ctrl + Alt + M`.  

1.  __Stage files to be included in the next commit__.  To stage a single 
    file, tick the corresponding check box. To stage all files, press Cmd + A, 
    then click `r bookdown::embed_png("screenshots/git-stage.png", dpi = 220)`. 
    
    As you stage each file, you'll notice that its status changes. There are two 
    columns: staged status on the left and unstaged status on the right. All
    staged change will be included in the next commit.
    
    There are two new new statuses that you'll see when staging files:

      * Added: `r bookdown::embed_png("screenshots/git-added.png", dpi = 220)`: 
        after staging an untracked file, git now knows that you want to add it 
        to the repo.
      
      * Renamed: `r bookdown::embed_png("screenshots/git-renamed.png", dpi = 220)`:
        If you rename a file, git initially sees it as deleting one file and
        adding a new file. Once you stage both changes, git will recognise 
        that it's a rename.
        
    Sometimes you'll see a status in both columns, e.g. 
    `r bookdown::embed_png("screenshots/git-modified-staged.png", dpi = 220)`.
    This means that you have both staged unstaged changes in the same file.
    This means that you've made some changes, staged them, and then made some 
    more. Clicking the staged checkbox will stage your new changes, clicking
    it again will unstage both sets of changes.

1.  __Write a commit message__ (top-right panel). The first line of a commit 
    message is called the subject line and should be brief (50 characters or 
    less). For complicated commits, you can follow it with a blank line and 
    then a paragraph or bulleted list providing more detail. Write messages in 
    imperative, like you're telling someone what to do: "fix this bug", not 
    "fixed this bug".
    
1.  __Click Commit__.

### Ignoring files

There are often files that you don't want to include in the repository. They might be transient artefacts (like stuff you get when building latex files, or compiling C code), or they might be too big, or they might be generated on demand. Instead of carefully not-staging them each time, you should add them to `.gitignore` to prevent them from every being added. 

The easiest way to do this is to right-click on the file in the git pane and select `Ignore`. If you want to ignore multiple files, you can use a wildcard "glob" like `*.png`.

Some developers stick to the convention that no files that can be generated from other files should ever be committed into git. For R packages this means the `man/` directory that is autogenerated from R comments. Pragmatically, it's better not to do this because currently R packages have no way to generate `.Rd` files on install, so if you don't check them in and someone installs your package from github they won't have any documentation.

## Commit best practices {#commit-best-practices}

Ideally, each commit should:

* __Related__: All the changes in the commit should be related to the same
  goal. This makes it easier to understand the commit at a glance, and to 
  describe it with a simple message.

* __Minimal__: A commit should contain nothing but the related changes. If 
  while while making one set of changes you discovered another problem,
  do a separate commit.

* __Complete__: A commit should do solve the problem that it claims to solve.
  If you think you've fixed a bug, the commit should contain a unit test
  that confirms that you're right.

And each commit message should:

* __Be concise, yet evocative__. You want to be able to take in at a glance
  what a commit does, but there should be enough detail so you can remember
  (or understand) what happened.

* __Describe the why, not the what__. You can always retrieve the diff 
  associated with a commit, so the message doesn't need to say exactly what 
  changed. Instead it should provide a high-level summary, and focus on the 
  motivation for the change.

If you do this then:

* It's easier to work with others. If two people have changed the same file
  in the same place, it's easier to resolve the conflict if the commits are
  small and it's clear why each change was made.
  
* Project newcomers can easily understand history by reading the commit logs.
  
* You can load and run your package at any point in its life. This can
  be tremendously useful in conjunction with tools like 
  [bisectr](https://github.com/wch/bisectr), which allow you to use binary
  search to quickly narrow in on the commit that introduced a bug.

* When you discover exactly when you introduced a bug, you can easily 
  understand what you were doing (and why!).

The time you spend on your commits is inversely proportional to the amount of time others will spend understanding the history of your project. This is a less common operation, but it's important if you want to see what's changed or undo a mistake.  

Commit messages are most important when you are collaborating with others. You might think that no one else will ever look at your repo, but there is one important collaborator for every project that you do: future-you! Future-you probably has forgotten a lot of things that present-you knows.

Remember that these directives are aspirational, and you shouldn't let aspirations get in your way.  If you look at the commit history of my repositories, you'll notice a lot of them aren't that great, especially when I start to get frustrated that I __still__ haven't managed to fix a bug. I'm a frequent offender on the "complete" front, often missing important parts of a problem.

## Synchronising with github {#github-init}

Most of the time you work locally. This is really convenient because you don't need internet access to use git - you can keep on committing changes even when you're on a plane. However, git is most useful in conjunction with github, allowing you to share your code with the world.  

To push your code to github:

1.  Create a new repo on github: <https://github.com/new>.

1.  In the shell, follow the instructions provided by github on the new
    repo page. They'll look something like this:

    ```bash
    git remote add origin git@github.com:rstudio/ggcomp.git
    git push -u origin master
    ```
  
    This tells git where the "origin" of your repo is, and to send it all the
    changes you've made so far.
    
    You only need to run these commands once per repo.
  
1.  Add `URL` and `BugReports` fields to your `DESCRIPTION` to link to your 
    new github site. For example, dplyr has:

    ```yaml
    URL: http://github.com/hadley/dplyr
    BugReports: http://github.com/hadley/dplyr/issues
    ```

Now you can __push__ your changes to github by clicking `r bookdown::embed_png("screenshots/git-push.png", dpi = 220)`. (This is the same as running `git push` in the shell). 

Typically, each push will include many commits. Pushing your code to github publishes it for the world to use. For that reason, once you've learned about `R CMD check` in [automated checking](#check), it's a good idea to run that before you push. Fix any problems so that you're always publishing good code.

Once you've connected your repo to github, the git pane will show you how many commits you have locally that are not on github: `r bookdown::embed_png("screenshots/git-local-commits.png", dpi = 220)`. This message indicates that I have 1 commit locally (my branch) that is not on github ("origin/master").

## Github 

The first benefit to using git is that you get a decent website autogenerated from your github commits. The home page, e.g. <https://github.com/hadley/testthat> (the github repo for testthat), lists all the files and directories in the top-level. You can navigate throughout the project and look at the files. Any `.md` files well be automatically rendered as html. If you have a `README.md` file in the top-level directory, it will be displayed when on the homepage. You'll learn more about the benefits of creating this file in [README.md](#readme).

The second benefit is that people can now easily install your package with `devtools::install_github("<username>/<repo>")`. 

## History {#git-log}

If you click on a file you'll see a nicely formatted version. This is useful for newcomers to your package - they can easily browse around and see what you package does before they download and try it out. There are three useful actions you can do to each file: `r bookdown::embed_png("screenshots/github-comment-line.png", dpi = 220)`:
  
  * Raw: this shows you the raw contents of the file, and can be useful if you
    want to copy-and-paste.
    
  * History: this lists every commit that has affected this particular file.
    This page is one reason its worth spending the time to write a good commit 
    message. If you have, you can skim this page to see recent changes to the 
    repo. 
    
    You can jump to this page directly from RStudio by clicking
    `r bookdown::embed_png("screenshots/github-view.png", dpi = 220)` in
    the git dropdown. If you have lines of code selected it will jump directly
    to those lines.
  
  * Blame: this is an alternative view of the history of the file. For each 
    line of code, it shows you the last commit to touch that line. This is 
    tremendously helpful when you've discovered a bug and want to understand
    its history.
    
    (It's called blame since it also shows you who made the commit so you
    know who to blame ;)
    
    You can jump to this page directly from RStudio by clicking
    `r bookdown::embed_png("screenshots/github-blame.png", dpi = 220)` in
    the git dropdown. 

You can also see every commit to your package in the commit view, e.g. <https://github.com/hadley/testthat/commits/master>. When I'm working on a package with other people, I often keep this page open so I can see what they're working on.  Individual commits show the same information that you see in the commit/diff window in RStudio, although github has some more advanced features that highlight individual words that have changed in sentences, and you can choose to show changes side-by-side instead of interleave in one file (a style called unified).

In this view, you can comment on the commit as a whole by using the comment box at the bottom of the page, or more usefully you can comment on individual lines by clicking `r bookdown::embed_png("screenshots/github-comment-line.png", dpi = 220)`. This is a great way to let your collaborators know if you see a mistake or have a question. It's better than email because it's public so anyone working on the repo (both present and future) can see the conversation.

## Undo a mistake {#git-undo}

The best thing about using commits is that you can undo mistakes. RStudio provides some tools for the most common types of mistake:

* If you want undo the changes you've made to a file, right click on it in 
  the git pane and select "revert". Proceed with caution: this is one of the
  few operations that you can't undo.

* You undo changes to parts of a file in the diff window. Look for the
  discard chunk button above the block of changes that you want to undo:
  `r bookdown::embed_png("screenshots/git-chunk.png", dpi = 220)`

* If you made a mistake in a commit, you can modify the previous commit by 
  clicking `r bookdown::embed_png("screenshots/git-commit-amend.png", dpi = 220)`. 
  (But don't do this if you've pushed the previous commit to github - you're
  effectively rewriting history, which should be done with care when you're 
  doing it in public.)
  
If one of these doesn't help, don't panic. The chances are that you can recover whatever was lost. Start by making a backup of the directory. Then even if you make a mistake while trying to fix your first mistake, you can start again. To undo more general problems you'll need to use the command line. 

The first step in undoing a mistake is to go back in time and find a good commit. The easiest way to do that is to use the "History" view to find the commit where the mistake occured. Then the __parent__ of that commit will be good. Note the sha (the unique identifier for that commit) of the parent. Then you can:

  * See what the file looked in the paste so you can copy-and paste the old
    code:
    
    ```bash
    git show <sha> <filename>
    ```

  * Copy historical version into the present:

    ```bash
    git checkout <sha> <filename>
    ```
    
In both cases you'll need finish by staging and commiting the files.

(It's also possible to use git as if you went back in time and prevented the mistake from happening in the first place. This is called __rebasing history__ and is an advanced technique. As you can imaging going back in time to change the past can have profound impacts on the present. It can be useful, but needs extreme care.)

If you're still stuck, try <http://sethrobertson.github.io/GitFixUm/fixup.html> or <http://justinhileman.info/article/git-pretty/>. They give step-by-step approach to fixing many common (and not so common!) problems.

## Working with others {#git-pull}

You use __push__ to send your changes to github. If you're working with other people, they're also pushing their changes to github, so you'll need to __pull__ their changes from github to see them locally. (You can only push to a repo when you have the most recent version of it, so you need to pull before you push your changes).

When you pull, git first downloads (__fetches__) all of the changes then __merges__ them with the changes that you've made. A merge is a commit with two parents. It takes two different lines of development and combines them into a single result. To do so, git needs to combine two sets of changes. In many cases it can do so automatically - if the changes are to different files, or even different parts of the same file. However, if the changes are to the same place in a file, you'll need to resolve the changes yourself. This is called a __merge conflict__.

In RStudio, you'll discover that you have merge conflict when:

* Pull fails with an error.

* In the git pane, you see a status like
  `r bookdown::embed_png("screenshots/git-commit-conflict.png", dpi = 220)`

RStudio currently doesn't provide any tools to help with merge conflicts, so you'll need to use the command line. I recommend starting by setting your merge conflict "style" to diff3. The diff3 style shows three things when you get a merge conflict: your local changes, the original file and remote changes. The default style is diff2, which only shows your changes and the remote changes, which generally makes it harder to figure out what's happened. 

* If you've encountered your first merge conflict:

    ```bash
    # Abort this merge
    git merge --abort
    # Set the conflict style
    git config --global merge.conflictstyle diff3
    # Re-try the merge
    git pull
    ```
    
* If you're not in the middle of a merge conflict, just run 

    ```bash
    git config --global merge.conflictstyle diff3
    ```

To resolve the merge conflict, you need to open every file with the status `r bookdown::embed_png("screenshots/git-commit-conflict.png", dpi = 220)`. In each file, you'll find a conflict marker that looks like this:

```
<<<<<<< HEAD

||||||| merged common ancestors

=======

>>>>>>> remote
```

This shows all three versions of the conflicting code:

* At top, the code that you code.
* In the middle, the code from the previous commit. (This is missing in the 
  default merge conflict style, so if you don't see it, follow the instructions
  above).
* At bottom, the code that you're trying to merge in.

You need to work through each conflict and decide either which version is better, or how combine the changes from both parents. Make sure you delete the conflict markers to resolve the conflict, then stage the file. Once you've fixed all conficts, make a new commit and push to github.

A couple of pointers when fixing text generated by roxygen:

* Don't fix problems in `man/*.Rd` files. Instead, resolve the conflicts in 
  the underlying roxygen comments and re-document the package.
   
* Merge conflicts in the `NAMESPACE` file will prevent you from re-loading or 
  re-documenting the package. Resolve them enough so that the package 
  can be loaded, then re-document to generate a clean and correct `NAMESPACE`.

Handling merge conflicts is one of the trickiest parts of git, and you made need to read a few tutorials before you get the hang of it. Google and StackOverflow are great resources. If you get terribly confused, you can always abort the merge and try again by running `git merge --abort` then `git pull`.

## Issues {#github-issues}

Every github repo comes with a page for tracking issues. Use them!  If you encounter a bug while working on another project, jot a note down in your issues. When you have a smaller project, don't worry too much about milestones, tags and assigning issues to specific people. Those are more useful once you get over one page of issues (~50). Once you get to that point, read the github guide on issues: <https://guides.github.com/features/issues/>.

One particularly useful tip is that you can close issues with commit messages. Just put `Closes #<issue number>` somewhere in your commit message and github will close the issue for you when you next push. The best thing about closing issues that way is that it makes a link from the issue to the commit. This is useful if you ever have to come back to the bug and see exactly what you did to fix it. If you want to make the link without closing the issue, you can just do `#<issue number>`.

Whenever you close an issue, it's a good idea to get in to the habit of add a bullet point to `NEWS.md`. See [NEWS.md](#news) for more details. The bullet point should describe the issue in terms that users will understand, where commit messages are designed to be read by developers.

## Branches {#git-branch}

Sometimes you need to make a big change to your code. You want to break it up into multiple simple commits so you can easily track what you're doing, but you don't want to merge them into the main stream of development. Or maybe you're not sure what you've done is the best approach and you want someone else to review your code. Or maybe you want to try something experimental - you only want to merge it back in if the experiment succeeds.  Branches and pull requests provide powerful tools to handle these situations.

Although you haven't realised it, you're already using branches. The default branch is called __master__, and it's where you've been saving your commits. If you synchronise your code to github, you have another branch called __origin/master__, which is a local copy of all the commits on github (synchronised whenever you push or pull). When you run `git pull`, it does two things:

* `git fetch origin master`, to downloads all the commits on github into the 
  `origin/master`
  
* `git merge orgin/master`, to combine the remote changes with your changes.

It's useful to create your own new branches when you want to break off the main stream of development for a while. Create a new branch with `git checkout -b <branch-name>`. Names should use lower case letters, numbers, and `-` to separate words
  
Switch between branches with `git checkout <branch-name>`. For example, to return to the main line of development with `git branch master`. You can also use the branch switcher at the top right of the git pane:
  
```{r, echo = FALSE}
bookdown::embed_png("screenshots/git-branch.png", dpi = 220)
```

If you've forgotten the name of your branch in the shell, you can use `git branch` to list all existing branches. 

If you try to synchronise this branch to github from inside RStudio, you'll notice that push and pull are disabled: `r bookdown::embed_png("screenshots/git-no-remote.png", dpi = 220)`. To enable them, you'll need to first tell git that your local branch should have a remote equivalent:

```bash
git push --set-upstream origin <branch-name>
```

Once you've done that the first time you can use the pull and push buttons like usual.


If you've been working in a branch for a while, other work might have been going on in master. To integrate that work into your branch, run `git merge master`. You will need to resolve any merge conflicts as described above. It's best to do this fairly frequently - the less your branch has diverged from the master, the easier it will be to merge.

Once you're done working in a branch, merge it back into master, then delete it:

```bash
git checkout master
git merge <branch-name>
git branch -d <branch-name>
```

(Git won't let you delete a branch unless you've merged it back into master. If you do want to abandon a branch without merging it, you'll need to force delete it by using `-D` instead of `-d`.)

## Making a pull request

Instead of merging your changes directly back into the main branch, you might want to first get someone else to review them. A __pull request__ allow to do just that. A pull request is a github tool that allows you to review a set of changes. A pull request has three components:

* A __conversation__, 
  `r bookdown::embed_png("screenshots/pr-conversation.png", dpi = 220)`,
  where you can discuss the changes as a whole.
  
* The __commits__, 
  `r bookdown::embed_png("screenshots/pr-commits.png", dpi = 220)`,
  where you can see each individual commit.
  
* The __file changes__,
  `r bookdown::embed_png("screenshots/pr-changes.png", dpi = 220)`,
  where you see the overall diff of the commits, and you can comment
  on individual lines.

We'll talk about the most common usage of pull requests, contributing to other people's code, next. But it's easiest to start using pull requests with your own repo because you own all the pieces, and it's easy to experiment. Pull requests against your own code are particularly useful when you want to get feedback on a proposed change.

To create a pull request, you need to work in a branch and synchronise the branch with github. Then when you next go to github you'll see a header that invites you to submit a pull request.
<https://help.github.com/articles/using-pull-requests/>

You can also do it by:

* switching branches: 

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/github-branches.png", dpi = 220)
    ```

* clicking `r bookdown::embed_png("screenshots/pr-create.png", dpi = 220)`

## Reviewing and accepting pull requests {#pr-accept}

<http://sarah.thesharps.us/2014/09/01/the-gentle-art-of-patch-review/>

Add bullet to news. 

## Making a pull request on other repos {#pr-make}

Making a pull request to someone else's code is similar to yours:

Adds another layer. You have a remote repo with the changes, and then you request to incorporate the changes into another. A pull request is a branch, so if you change the branch you'll also change the pull request. This is a feature because it allows you to update the pull request in response to comments from the maintainer. But it means that you should always create a branch for each pull request.

1.  Fork the repo. A fork is a copy of the repo that you can work on 
    independently. See [fork a repo](https://help.github.com/articles/fork-a-repo) 
    for more details.

1.  Creating a local copy of a remote repo is called __cloning__. You can do it
    from the shell with `git clone`, or use RStudio. Go to the github page and 
    determine the git repo url. You need to clone your forked repo, not the 
    original (otherwise you can't push any changes to github). It's shown in 
    the right-hand sidebar:

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-clone-github.png", dpi = 220)
    ```
    
    I recommend using the ssh url (by clicking SSH). Click the copy button
    to put the url on your clipboard.
    
1.  Create a new project in RStudio:

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-clone-1.png", dpi = 220)
    bookdown::embed_png("screenshots/git-clone-2.png", dpi = 220)
    ```
    
1.  Paste the repo url:

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/git-clone-3.png", dpi = 220)
    ```

    (Here I'm cloning the repo for this book.)

1.  Tell the fork about the original repo:

    ```bash
    git remote add upstream git@github.com:wch/extrafont.git
    git fetch upstream
    git branch -u upstream/master
    ```

Good commit messages are most important when you're submitting a pull request, sharing your changes with someone else. You want to make life as easy as possible for them, so that they can easily see what you've done, decreasing the amount of work they have to do and increasing the chances that your pull request will be accepted.

* Strive to make as clean as possible.
* Branch

There are more advanced history modifying features available from the command
line. See [rewriting history](http://git-scm.com/book/en/v2/Git-Tools-Rewriting-History)
for details.

