---
layout: default
title: Introduction
output: bookdown::html_chapter
---

# Introduction

## Why build a package?

In R, the fundamental unit of shareable code is the package. A package bundles together code, data, documentation and tests, and it's easy to share with others. A package is just a set of conventions. Each convention is an instruction for organising your code, and each chapter of this book describes one convention. For example, in the first two chapter you'll learn to:

* Put your R code in a directory called `R/`.

* Describe what your package needs to work in a `DESCRIPTION` file
  If the package is to share, also describe what it does, who can use it 
  (the license) and who to contact if things go wrong.

In later chapters, you'll learn about:

* function documentation, which lives in `man/`.

* vignettes, long-form tutorials, which live in `vignettes/`.

* the `NAMESPACE` file which describes how you package connects to other
  packages.
  
* and much much more.

These conventions are helpful because:

* They save you time --- you don't need to think about the best way to organise
  a project, you can just follow a template.

* Standardised conventions lead to standardised tools --- if you buy into
  the R's package conventions, you get many tools for free.

As Hilary Parker puts it in her [introduction to packages](http://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/): "Seriously, it doesn't have to be about sharing your code (although that is an added benefit!). It is about saving yourself time."

A package doesn't need to be complicated. You can start with a minimal subset of useful features and slowly build up over time. While there are strict requirements if you want to publish a package to the world on CRAN (and many of those requirements are useful even for your own packages), most packages won't end up on CRAN. You can even use packages to structure your data analyses, as Robert M Flight discusses in a [series of blog posts](http://rmflight.github.io/posts/2014/07/analyses_as_packages.html). Packages are really easy to create and use once you have the right set of tools.

Anytime you create some reusable set of functions you should put it in a package. It's the easiest path because packages come with conventions: you don't need to figure them out for yourself. You'll start with just your R code in the `R/` directory, and over time you can flesh it out with documentation (in `man/`), vignettes (in `vignettes/`), compiled code (in `src/`), data sets (in `data/`), and tests (in `inst/tests`).

The most accurate resource for up-to-date details on package development is always the official [writing R extensions][r-ext] guide. However, it's rather hard to understand if you're not already familiar with the basics of packages. It's also exhaustive, covering every possible package component, rather than focussing on the most common and useful components as this book does. Writing R extensions is a useful resource once you've mastered the basics and need to check on more esoteric facts.

## Philosophy

This book espouses my philosophy of package development: anything that can be automated, should be automated. Do as little as possible by hand. Do as much as possible with functions. The goal is to spend your time thinking about what you want your package to do rather than thinking about the minutiae of package structure.

This philosophy is realised primarily through the devtools package, a suite of R functions that automates common development tasks. The goal of devtools is to make package development as painless as possible. It does so by putting into a package of its own all the best practices of package development that I've learned over the years. With devtools, you won't have to worry or even know about these practices. And, as I discover and incorporate even better ways of doing things, all you'll need to do to reap the benefits is to simply upgrade devtools.

Developing packages with devtools may take some getting used to. However, I strongly believe that the time invested in mastering it will, in the long run, save you much, much more time. Devtools is a key component of my productivity in package development.

Devtools works hand-in-hand with RStudio, which I believe is the best development environment for most R users. The only real competitor is [ESS](http://ess.r-project.org/), emacs speaks statistics, which is a rewarding environment if you're willing to put in the time to learn emacs and customise it to your needs.

## Getting started

To get started, make sure you have the latest version of R (at least 3.1.0, which was current when I wrote this book), then run the following code to get the packages you'll need:

```{r, eval = FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

Make sure you have a recent version of RStudio. Since you're making the jump from an R user to a package developer, I recommend that you download the preview version from <http://www.rstudio.com/products/rstudio/download/preview/>. This will give you access the latest and greatest features, and will only slightly increases your chances of finding a bug.

During the development of the book, you'll probably also want the development version of devtools. This gives you access new devtools functions as I develop them:

```{r, eval = FALSE}
devtools::install_github("hadley/devtools")
```

You'll also need to make sure you have a C compiler and other needed command line tools. If you're on Windows or Mac and you don't already have them, RStudio will install them for you. Otherwise:

* On Windows, download and install [Rtools](http://cran.r-project.org/bin/windows/Rtools/). NB: this is not an R 
  package!

* On Mac, make sure you have either XCode (available for free in the App Store)
  or the ["Command Line Tools for Xcode"](http://developer.apple.com/downloads).
  NB: you'll need to have an Apple ID (available for free).

* On Linux, make sure you've installed not only R, but the R development
  devtools. This is a Linux package called something like `r-base-dev`.

You can check you have everything installed and working by running the following:

```{r, eval = FALSE}
library(devtools)
has_devel()
#> '/Library/Frameworks/R.framework/Resources/bin/R' --vanilla CMD SHLIB foo.c 
#> 
#> clang -I/Library/Frameworks/R.framework/Resources/include -DNDEBUG 
#>   -I/usr/local/include -I/usr/local/include/freetype2 -I/opt/X11/include
#>   -fPIC  -Wall -mtune=core2 -g -O2  -c foo.c -o foo.o
#> clang -dynamiclib -Wl,-headerpad_max_install_names -undefined dynamic_lookup
#>   -single_module -multiply_defined suppress -L/usr/local/lib -o foo.so foo.o 
#>   -F/Library/Frameworks/R.framework/.. -framework R -Wl,-framework 
#>   -Wl,CoreFoundation
[1] TRUE
```

This will print out some compilation code (this is needed to help diagnose problems). But you're really only interested in whether it returns `TRUE` (everything's OK) or an error (further investigation needed).

## Package basics

This book is arranged according to the structure of an R package. It starts with the most commonly used components, and then gets more esoteric.

* `R/`: where your R code lives in `.R` files

* `DESCRIPTION`: metadata about the package

* `man/`: function documentation

* `vignettes/`: long-form documentation which show how to combine multiple 
  parts of your package to solve real problems.

* `NAMESPACE`: ensures that your package plays nicely with other packages.

* `tests/`: stores unit tests that ensure that your package is operating as
  designed.

* `data/`: sample datasets (or other R objects)

* `src/`: compiled C, C++ and Fortran source code

## Acknowledgments {#intro-ack}

I would like to thank the tireless contributors to R-devel and, more recently, [stackoverflow](http://stackoverflow.com/questions/tagged/r). There are too many to name individually, but I'd particularly like to thank Dirk Eddelbuettel, Martin Morgan, Luke Tierney, and Brian Ripley for generously giving their time and correcting my countless misunderstandings.

This book was [written in the open](https://github.com/hadley/r-pkgs/), and chapters were advertised on [twitter](https://twitter.com/hadleywickham) when complete. It is truly a community effort: many people read drafts, fixed typos, suggested improvements, and contributed content. Without those contributors, the book wouldn't be nearly as good as it is, and I'm deeply grateful for their help. 

```{r, results = "asis", echo = FALSE, eval = FALSE}
#contribs <- system("git --no-pager shortlog -ns > contribs.txt", intern = T)
contribs <- read.delim("contribs.txt", header = FALSE, 
  stringsAsFactors = FALSE)[-1, ]
names(contribs) <- c("n", "name")

contribs <- contribs[order(contribs$name), ]
contribs$uname <- ifelse(!grepl(" ", contribs$name), 
  paste0("@", contribs$name), contribs$name)

cat("Thanks go to all contributers in alphabetical order: ")
cat(paste0(contribs$uname, collapse = ", "))
cat(".\n")
```

## Conventions

Throughout this book I use `f()` to refer to functions, `g` to refer to variables and function parameters, and `h/` to paths. 

Larger code blocks intermingle input and output. Output is commented so that if you have an electronic version of the book, e.g., <http://r-pkgs.had.co.nz>, you can easily copy and paste examples into R. Output comments look like `#>` to distinguish them from regular comments.

## Colophon

This book was written in [Rmarkdown](http://rmarkdown.rstudio.com/) inside [RStudio](http://www.rstudio.com/ide/). [knitr](http://yihui.name/knitr/) and [pandoc](http://johnmacfarlane.net/pandoc/) converted the raw Rmarkdown to html and pdf. The [website](http://adv-r.had.co.nz) was made with [jekyll](http://jekyllrb.com/), styled with [bootstrap](http://getbootstrap.com/), and automatically published to Amazon's [S3](http://aws.amazon.com/s3/) by [travis-ci](https://travis-ci.org/). The complete source is available from [github](https://github.com/hadley/r-pkgs).

This version of the book was built with:

```{r}
devtools::session_info()
```

[r-ext]:http://cran.r-project.org/doc/manuals/R-exts.html#Creating-R-packages
