---
layout: default
title: Introduction
output: bookdown::html_chapter
---

# Introduction

## Why build a package?

Packages make your life easier.

By bundling code, data, documentation and tests together into a single unit, packages make sharing your code easy. In R, the package is the fundamental unit of shareable code.

By defining a set of conventions about how to organise your code, packages help you save time (See Hilary Parker's [introduction to packages](http://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)). All you need to do is follow a template. The first convention is a no-brainer:

* Put your R code in a directory called R/.

The second is a little bit more work:

* Describe your package's requirements in a `DESCRIPTION` file. If the package will be shared, explain what it does, who can use it (the license) and whom to contact (if things go wrong).

Such conventions are helpful because:

* You don’t need to think about the how to best organise a project, you just follow a template.

* By following them, you get access to many tools for free.
    
A package needn't to be complicated. In fact, anytime you create a reusable set of functions you should put it in a package. Over time, you can add useful functions and features. To do that, you should follow the requirements for publishing packages on CRAN: beyond putting your R code in the R/ directory, and over time you can flesh it out with documentation (in man/), vignettes (in vignettes/), compiled code (in src/), data sets (in data/), and tests (in inst/tests). 

While these requirements are strict and even though most packages won’t end up there, they are are useful: for example, you can use packages to structure your data analyses, as Robert M Flight discusses in a [series of blog posts](http://rmflight.github.io/posts/2014/07/analyses_as_packages.html). More importantly, following these conventions will make creating packages easy.

The most accurate resource for up-to-date details on package development is the official [Writing R Extensions][r-ext] guide. However, if you're not already familiar with the basics of packages it's rather hard to understand. This is because rather than focussing on the most commonly used components (as this book does), it's an exhaustive treatment which covers every possible aspect of writing packages. Once you've mastered the basics and need to check on more esoteric facts, Writing R Extensions will be a useful resource.

## Philosophy

This book espouses my philosophy of package development: anything that can be automated, should be automated. You will do as little as possible by hand, and instead rely functions to do as much work as possible. The goal is to let you spend your time thinking about what you want your package to do, not thinking about the minutiae of package structure.

This philosophy is realised primarily through the devtools package, which provides a suite of R functions to automate common development tasks. The goal of the devtools package is to make package development as painless as possible. Devtools takes what I know about package development best practices, and encodes them in functions so you don't have to remember, or even know about the best way to tackle a problem. As I discover better ways of doing things, all you need to do to reap the benefits is to upgrade devtools.

If you've developed packages without devtools, it may take a bit of getting used to. However, I strongly believe that mastering it will save you much much more time than it will take to learn. Devtools is a key component of my package development productivity.

Devtools works hand-in-hand with RStudio, which I believe is the best development environment for most R users. The only real competitor is [ESS](http://ess.r-project.org/), emacs speaks statistics, which is a rewarding environment if you're willing to put in the time to learn emacs and customise it to your needs.

## Getting started

To get started, make sure you have the latest version of R (at least 3.1.0, which was current when I wrote this book), then run the following code to get the packages you'll need:

```{r, eval = FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

Make sure you have a recent version of RStudio. Since you're making the jump from an R user to a package developer, I recommend that you download the preview version from <http://www.rstudio.com/products/rstudio/download/preview/>. This will let you access the latest and greatest features, and only slightly increases your chances of finding a bug.

During the development of the book, you'll probably also want the development version of devtools. This will let you access new devtools functions as I develop them:

```{r, eval = FALSE}
devtools::install_github("hadley/devtools")
```

You'll also need to make sure you have a C compiler and other need command line tools. If you're on windows or the mac, Rstudio will install these for you if you don't have them already. Otherwise:

* On Windows, download and install [Rtools](http://cran.r-project.org/bin/windows/Rtools/). NB: this is not an R package!

* On Mac, make sure you have either XCode (free, available in the app store)
  or the ["Command Line Tools for Xcode"](http://developer.apple.com/downloads)
  (needs a free apple id).

* On Linux, make sure you've installed not only R, but the R development
  devtools. This is a Linux package called something like `r-base-dev`.

You can check you have everything installed and working by running this code:

```{r, eval = FALSE}
library(devtools)
has_devel()
#> '/Library/Frameworks/R.framework/Resources/bin/R' --vanilla CMD SHLIB foo.c 
#> 
#> clang -I/Library/Frameworks/R.framework/Resources/include -DNDEBUG 
#>   -I/usr/local/include -I/usr/local/include/freetype2 -I/opt/X11/include
#>   -fPIC  -Wall -mtune=core2 -g -O2  -c foo.c -o foo.o
#> clang -dynamiclib -Wl,-headerpad_max_install_names -undefined dynamic_lookup
#>   -single_module -multiply_defined suppress -L/usr/local/lib -o foo.so foo.o 
#>   -F/Library/Frameworks/R.framework/.. -framework R -Wl,-framework 
#>   -Wl,CoreFoundation
[1] TRUE
```

It will print out some compilation code (this is needed to help diagnose problems), but you're only interested in whether it returns `TRUE` (everything's ok) or an error (which you need to investigate further).

## Package basics

This book is arranged according to the structure of an R package. It starts with the most commonly used components, and then gets more esoteric.

* `R/`: where your R code lives in `.R` files.

* `DESCRIPTION`: metadata about the package

* `man/`: function documentation

* `vignettes/`: long-form documentation which show how to combine multiple 
  parts of your package to solve real problems.

* `NAMESPACE`: ensures that your package plays nicely with others.

* `tests/`: stores unit tests that ensure that your package is operating as
  designed.

* `data/`: sample datasets (or other R objects)

* `src/`: compiled C, C++ and fortran source code

## Conventions

Throughout this book I use `f()` to refer to functions, `g` to refer to variables and function parameters, and `h/` to paths. 

Larger code blocks intermingle input and output. Output is commented so that if you have an electronic version of the book, e.g., <http://r-pkgs.had.co.nz>, you can easily copy and paste examples into R. Output comments look like `#>` to distinguish them from regular comments.

## Colophon

This book was written in [Rmarkdown](http://rmarkdown.rstudio.com/) inside [RStudio](http://www.rstudio.com/ide/). [knitr](http://yihui.name/knitr/) and [pandoc](http://johnmacfarlane.net/pandoc/) converted the raw Rmarkdown to html and pdf. The [website](http://adv-r.had.co.nz) was made with [jekyll](http://jekyllrb.com/), styled with [bootstrap](http://getbootstrap.com/), and automatically published to Amazon's [S3](http://aws.amazon.com/s3/) by [travis-ci](https://travis-ci.org/). The complete source is available from [github](https://github.com/hadley/r-pkgs).

```{r}
devtools::session_info()
```

[r-ext]:http://cran.r-project.org/doc/manuals/R-exts.html#Creating-R-packages
