# Lifecycle

## Version {#description-version}

Formally, an R package version is a sequence of at least two integers separated by either `.` or `-`.
For example, `1.0` and `0.9.1-10` are valid versions, but `1` and `1.0-devel` are not.
You can parse a version number with `numeric_version()`.

```{r}
numeric_version("1.9") == numeric_version("1.9.0")
numeric_version("1.9.0") < numeric_version("1.10.0")
```

For example, a package might have a version 1.9.
This version number is considered by R to be the same as 1.9.0, less than version 1.9.2, and all of these are less than version 1.10 (which is version "one point ten", not "one point one zero").
R uses version numbers to determine whether package dependencies are satisfied.
A package might, for example, import package `devtools (>= 1.9.2)`, in which case version 1.9 or 1.9.0 wouldn't work.

Here is our recommended framework for managing the package version number:

* Always use `.` as the separator, never `-`.

* A released version number consists of three numbers, `<major>.<minor>.<patch>`. 
  For version number 1.9.2, 1 is the major number, 9 is the minor number, and 
  2 is the patch number.
  Never use versions like `1.0`, instead always spell out the three components,
  `1.0.0`.
  
* An in-development package has a fourth component: the development version.
  This should start at 9000.
  For example, the first version of the package should be `0.0.0.9000`.
  There are two reasons for this recommendation:
  First, it makes it easy to see if a package is released or in-development.
  Also, the use of the fourth place means that you're not limited to what the
  next version will be.
  `0.0.1`, `0.1.0`, and `1.0.0` are all greater than `0.0.0.9000`.
  
  Increment the development version, e.g. from `9000` to `9001`, if you've
  added an important feature that another development package needs to depend 
  on.

The advice above is inspired in part by [Semantic Versioning](https://semver.org) and by the [X.Org](https://www.x.org/releases/X11R7.7/doc/xorg-docs/Versions.html) versioning schemes.
Read them if you'd like to understand more about the standards of versioning used by many open source projects.
Finally, know that other maintainers follow different philosophies on how to manage the package version number.

The version number of your package increases with subsequent releases of a package, but it's more than just an incrementing counter -- the way the number changes with each release can convey information about what kind of changes are in the package.
We discuss this and more in Section \@ref(release-version).
For now, just remember that the first version of your package should be `0.0.0.9000`.
`usethis::create_package()` does this, by default.
`usethis::use_version()` increments the package version; when called interactively, with no argument, it presents a helpful menu:

```{r, eval = FALSE}
usethis::use_version()
#> Current version is 0.1.
#> What should the new version be? (0 to exit) 
#> 
#> 1: major --> 1.0
#> 2: minor --> 0.2
#> 3: patch --> 0.1.1
#> 4:   dev --> 0.1.0.9000
#> 
#> Selection: 
```