# Other documentation

```{r, echo = FALSE}
source("common.R")
status("drafting")
```

## Introduction

## Documenting packages {#man-packages}

You can use roxygen to provide a help page for your package as a whole.
This is accessed with `package?foo`, and can be used to describe the most important components of your package.
It's a useful supplement to vignettes, as described in the next chapter.

There's no object that corresponds to a package, so you need to document `NULL`, and then manually label it with `@docType package` and `@name <package-name>`.
This is also an excellent place to use the `@section` tag to divide up page into useful categories.

```{r}
#' foo: A package for computating the notorious bar statistic
#'
#' The foo package provides three categories of important functions:
#' foo, bar and baz.
#' 
#' @section Foo functions:
#' The foo functions ...
#'
#' @docType package
#' @name foo
NULL
```

I usually put this documentation in a file called `<package-name>.R`.
It's also a good place to put the package level import statements that you'll learn about in @sec-imports.
