---
title: Namespaces
layout: default
output: bookdown::html_chapter
---

# Namespace {#namespace}

The package `NAMESPACE` is one of the most confusing parts of building a package. Roxygen2 aims to make it as easy as possible to build a package that is a well-behaved member of the R ecosystem. This is a little frustrating at first, but soon becomes second-nature.

As the name suggest, namespaces provide "spaces" for "names", providing a context for evaluating which object is found when you look for it. When writing code, they allow you to specify which package to look for a function when there are multiple packages it could come from, and when developing packages they ensure that your functions always call functions in the same place.

Namespaces make it possible for packages to refer to specific functions in other packages, not the functions that you have defined in the global workspace.

For example, take the simple `nrow()` function:

```{r}
nrow
```

It's defined in terms of `dim()`, so what happens if we override `dim()` with our own definition --- does `nrow()` break?

```{r}
dim <- function(x) c(1, 1)
dim(mtcars)
nrow(mtcars)
```

Suprisingly, it does not! That's because when `nrow()` looks for an object called `dim()`, it finds the function in the base package, not our function.

Namespaces also provide an extension to the standard way of looking up the value of an object: instead of just typing `objectname`, you type `package::objectname`.  For example, both `plyr` and `Hmisc` provide a `summarise()` function. If you load `plyr`, then `Hmisc`, `summarise` will refer to the HMisc implementation; if you load the packages in the opposite order, then `summarise` will refer to the plyr implementation. This is obviously confusing, so instead, you can be explicit and refer to either `Hmisc::summarise` or `plyr::summarise` - then it doesn't matter in which order the packages are loaded.

Namespaces also control the functions and methods that your package makes available for use by others. Namespaces make it easier to come up with your own function names without worrying about what names other packages have used. A namespace means you can use any name you like for internal functions, and when there is a conflict with an exported function, there is a standard disambiguation procedure.

The easiest way to use namespaces is with roxygen2, because it keeps the namespace definitions next to the function that it concerns. The translation between roxygen2 tags and `NAMESPACE` directives is usually straightforward: the tag becomes a function name and its space-separated arguments become comma separated arguments to the function.  For example `@import plyr` becomes `import(plyr)` and `@importFrom plyr ddply` becomes `importFrom(plyr, ddply)`.  The chief exception is `@export` which will automatically figure out the function name to export. Only unique directives are saved to the `NAMESPACE` file, so you can repeat them as needed to maintain a close link between the functions where they are needed and the namespace file.

## Workflow

The workflow is very similar to creating `man/` files:

1. Add roxygen comments to your `.R` files.

1. Run `devtools::document()` (or press Cmd + Shift + D in RStudio) to 
   convert roxygen comments to `.Rd` files.

1. Look at `NAMESPACE` and run tests.

1. Rinse and repeat until the correct functions are exported.

## Exports

For a function to be usable outside of your package, you must __export__ it. By default roxygen2 doesn't export anything from your package. If you want an object to be publically available, you must explicitly tag it with `@export`.

Use the following guidelines to decide what to export:

* Functions: export functions that you want to make available. Exported
  functions must be documented, and you must be cautious when changing their
  interface. Note that all functions used in examples must be exported.

* Datasets: all datasets are publicly available. They exist outside of the
  package namespace and should not be exported.

### S3 

* S3 classes: if you want others to be able to create instances of the class
  `@export` the constructor function.

* S3 generics: the generic is a function so `@export` if you want it to
  be usable outside the package

* S3 methods: I recommend that you export every S3 method, even if the
  generic is not exported. Strictly speaking, you don't need to export methods
  for non-exported generics, but exporting every S3 method is a simpler rule
  and makes it less likely that you introduce hard to pin-down bugs. 
  You can use `devtools::missing_s3()` to identify list all S3 methods 
  that you've forgotten to export.

Things are a little more complicated if you want to define a method for a generic that you didn't create. If the method is in a required package (i.e., it's listed in Depends or Imports in the DESCRIPTION), then you need to import the generic into your package (see below). If the method is in a suggested package, you need to register S3 methods at run-time, when the other package is loaded.

From htmltools:

```{r}
# COPYRIGHT RStudio, GPL >= 2
registerMethods <- function(methods) {
  lapply(methods, function(method) {
    pkg <- method[[1]]
    generic <- method[[2]]
    class <- method[[3]]
    func <- get(paste(generic, class, sep="."))
    if (pkg %in% loadedNamespaces()) {
      registerS3method(generic, class, func, envir = asNamespace(pkg))
    }
    setHook(
      packageEvent(pkg, "onLoad"),
      function(...) {
        registerS3method(generic, class, func, envir = asNamespace(pkg))
      }
    )
  })
}
```

### S4 & RC

  * S4 classes: if you want others to be able to extend your class, `@export` it.
    If you want others to create instances of your class, but not extend it,
    `@export` the constructor function, but not the class.

    ```{r, eval = FALSE}
    # Can extend and create
    #' @export
    setClass("A")

    # Can extend, but constructor not exported
    #' @export
    B <- setClass("B")

    # Can create, but not extend
    #' @export C
    C <- setClass("C")

    # Can create and extend
    #' @export D
    #' @exportClass D
    D <- setClass("D")
    ```

* S4 generics: `@export` if you want the generic to be publicly usable.

* S4 methods: you only need to `@export` methods for generics that you
  did not define. But `@export`ing every method is a good idea as it
  will not cause problems and prevents you from forgetting to export an
  important method.

* RC classes: the same principles apply as for S4 classes. Note that due
  to the way that RC is implemented, it's typically impossible for your
  classes to be extended outside of your package.

## Imports {#imports}

The `NAMESPACE` also controls which functions from other packages are made available to your package. 

### Functions

If you are using just a few functions from another package, the recommended option is to note the package name in the `Imports:` field of the `DESCRIPTION` file and call the function(s) explicitly using `::`, e.g., `pkg::fun()`.  Alternatively, though no longer recommended due to its poorer readability, use `@importFrom`, e.g., `@importFrom pgk fun`, and call the function(s) without `::`.

If you are using many functions from another package, use `@import package` to import them all and make available without using `::`.

### S3

If you want to a method to an S3 defined in another package, import it with `@importFrom pkg generic`.

You don't need to do anything to access S3 methods defined in another package.

### S4 {#namespace-s4}

* To use classes defined in another package: 
  `@importClassesFrom package ClassA ClassB ...`. Place this next to the 
  class that inherit from the imported classes, or the methods that implement
  a generic for the imported classes.

* To use methods defined in another package: 
  `@importMethodsFrom package GenericA GenericB ...`. Place these next to the
  methods that use the imported generics.

You also need to make sure that the functions in the `methods` package are available to your code:

  * Pre R 3.2.0: `Depends: methods` in `DESCRIPTION`.  \
    Post R 3.2.0: `Imports: methods` in `DESCRIPTION`.
  
  * Since you'll being using a lot of functions from `methods`, 
    you'll probably also want to import the complete package with:
    
    ```{r}
    #' @imports methods
    NULL
    ```
    
    Or you might just wanted import the most commonly used functions:
    
    ```{r}
    #' @importFrom methods setClass setGeneric setMethod setRefClass
    NULL
    ```
  
    It doesn't matter where these import definitions go, but if you have
    package level docs, that's a natural place to put them.


### Compiled code

To import compiled code from another package, use `@useDynLib`

* `@useDynLib package` imports all compiled functions.

* `@useDynLib package routinea routineb` imports selected compiled functions.

* Any `@useDynLib` specification containing a comma, e.g.
  `@useDynLib mypackage, .registration = TRUE` will be inserted as is
  into the the `NAMESPACE`, e.g. `useDynLib(mypackage, .registration = TRUE)`
