---
layout: default
title: Package structure
output: bookdown::html_chapter
---

# Package structure {#package-structure}

This chapter will start you on the road to package development by showing you how to create your first package. You'll also learn about the various states a package can be in, including what happens when you install a package. Finally you'll learn about the difference between a package and a library and why you should care.

## Naming your package {#naming}

> "There are only two hard things in Computer Science: cache invalidation and 
> naming things." 
>
> --- Phil Karlton

Before you can create your first package, you need to come up with a name for it. I think this is the hardest part of creating a package!

### Requirements for a name

There's only one formal requirement: the name can only consist of letters, numbers and `.` (and it must start with a letter and cannot end with a period). Unfortunately this means you can't use `-` or `_` in your package name. I recommend against using `.` in package names because it has confusing connotations (i.e., file extension or S3 method).

### Strategies for creating a name

If you're planning on releasing your package, I think it's worth spending a few minutes to come up with a good name. Here are some recommendations for how to go about it:

* Pick a unique name so you can easily Google it. This makes it easy for 
  potential users to find your package (and associated resources), and it makes 
  it easier for you to see who's using it. You can check if a name is already
  used on CRAN by loading <http://cran.r-project.org/web/packages/[PACKAGE_NAME]>.

* Avoid using both upper and lower case letters: they make the package name
  hard to type and hard to remember. For example, I can never remember if it's
  Rgtk2 or RGTK2 or RGtk2.

* Find a word that evokes the problem and modify it so that it's unique:

    * plyr is generalisation of the apply family, and evokes pliers.
    * lubridate makes dates and times easier.
    * knitr (knit + r) is "neater" than sweave (s + weave).
    * testdat tests that data has the correct format.

* Use abbreviations: 

    * Rcpp = R + C++ (plus plus)
    * lvplot = letter value plots. 

* Add an extra R: 

    * stringr provides string tools.
    * tourr implements grand tours (a visualisation method).
    * gistr lets you programmatically create and modify github gists.

If you're creating a package that talks to a commercial service, make sure you check the branding guidelines to avoid problems down the line. For example, rDrop isn't called rDropbox because  Dropbox prohibits any applications from using the full trademarked name.

## Creating a package {#getting-started}

Once you've come up with a name, there are two ways to create the package. You can use Rstudio:

1. Click File | New Project.

2. Choose "New Directory":

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/create-1.png", dpi = 220)
    ```
   
2. Then "R Package":

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/create-2.png", dpi = 220)
    ```

2. Then give your package a name and click "Create Project":

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/create-3.png", dpi = 220)
    ```
    
Alternatively, you can create a new package from inside R by running

```{r, eval = FALSE}
devtools::create("path/to/package/pkgname")
```
    
Either route gets you to the same place: the smallest usable package, which has four components:

1. An RStudio project file, `pkgname.Rproj`, described below.

1. An `R/` directory, which you'll learn about in [R code](#r).

1. A basic `DESCRIPTION` file, which you'll learn about in 
   [package metadata](#description).

1. A basic `NAMESPACE` file, which you'll learn about in 
   [the namespace](#namespace).

__Never__ use `package.skeleton()` to create a package. It creates many additional files that you'll have to create or modify before you have a valid working package.

## RStudio projects {#projects}

To get started with your new package in RStudio, double-click the `pkgname.Rproj` file that `create()` just made. This will open a new RStudio project for your package. Projects are a great way to develop packages because:

*   Each project is isolated; unrelated things are kept unrelated.

*   You get handy code navigation tools like `F2` to jump to a function
    definition and `Ctrl + .` to look up functions by name.

*   You get useful keyboard shortcuts for common package development tasks.
    You'll learn about them throughout the book. But to see them all, press 
    Alt + Shift + K or use the Help | Keyboard shortcuts menu.
    
    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/keyboard-shortcuts.png")
    ```
   
(If you want to learn more RStudio tips and tricks, follow @[rstudiotips](https://twitter.com/rstudiotips) on twitter.)

Both Rstudio and `devtools::create()` will make an `.Rproj` file for you. If you have an existing package that doesn't have a `.Rproj` file, you can use `devtools::use_rstudio("path/to/package")` to add it. If you don't use RStudio, you can get many of the benefits by starting a new R session and ensuring the working directory is set to the package directory. 

### What is a RStudio project file?

`.Rproj` files are just text files. The project file created by devtools looks like this:

```
Version: 1.0

RestoreWorkspace: No
SaveWorkspace: No
AlwaysSaveHistory: Default

EnableCodeIndexing: Yes
Encoding: UTF-8

AutoAppendNewline: Yes
StripTrailingWhitespace: Yes

BuildType: Package
PackageUseDevtools: Yes
PackageInstallArgs: --no-multiarch --with-keep.source
PackageRoxygenize: rd,collate,namespace
```

Never modify this file by hand. Instead, use the friendly project options dialog, accessible from the projects menu in the top-right corner of RStudio.

```{r, echo = FALSE}
bookdown::embed_png("screenshots/project-options-1.png", dpi = 220)
bookdown::embed_png("screenshots/project-options-2.png", dpi = 220)
```

## What is a package? {#package}

To make your first package, all you need to know is what you've learnt above. To master package development, particularly when you're distributing a package to others,  The five states of an in-development package. you'll need to understand the differences between the five types of packages: source, bundled, binary, installed and in memory. Doing so will help you understand existing CRAN packages, and exactly what happens when you install a package with `install.packages()` or with `devtools::install_github()`.

### Source packages

So far we've just worked with a __source__ package: the development version of a package that lives on your computer. A source package is just a directory with components like `R/`, `DESCRIPTION`, and so on.

### Bundled packages

A __bundled__ package is a package compressed into a single file. By convention, package bundles in R use the extension `.tar.gz`. This convention comes from Linux: it means that multiple files have been collapsed into a single file (`.tar`) and then compressed using gzip (`.gz`). While a bundle is not that useful on its own, it is a useful intermediary between other steps. In the rare case that you do need a bundle, call `devtools::build()` to make it. 

If you decompress a bundle, you'll see it looks almost the same as your source package. The main differences between a decompressed bundle and a source package are:

* Vignettes are built so that you get html and pdf output instead of 
  markdown or latex input.

* Your source package might contain temporary files used to save time during
  development, like compilation artefacts in `src/`. These are never found in 
  a bundle.
  
* Any files listed in `.Rbuildignore` are not included in the bundle.

`.Rbuildignore` prevents files in the source package appearing in the bundled package. It allows you to have additional directories in your source package that are not included in the package bundle. This is particularly useful when you generate package contents (e.g. data) from other files. Those files should be included in the source package, but only the results need to be distributed. This is particularly important for CRAN packages (where the set of allowed top-level directories is fixed). Each line gives a Perl-compatible regular expression that is matched case-insensitively against the path to each file (i.e. `dir(full.names = TRUE)` run from the package root directory) - if the regular expression matches, the file is excluded. 

If you wish to exclude a specific file or directory (which is the most common use case), you __MUST__ escape the regular expression. For example, to exclude a directory called notes, use `$notes^`. The regular expression `notes` will match any file name containing notes, e.g. `R/notes.R`, `man/important-notes.R`, `data/endnotes.Rdata`, etc. The safest way to exclude a specific file or directory is to use `devtools::use_build_ignore("notes")`, which will do the escaping for you.

Here's a typical `.Rbuildignore` file from one of my packages:

```
^.*\.Rproj$         # Automatically added by RStudio,
^\.Rproj\.user$     # used for temporary files. 
^README\.Rmd$       # An Rmarkdown file used to generate README.md
^cran-comments\.md$ # Comments for CRAN submission
^NEWS\.md$          # A news file written in Markdown
^\.travis\.yml$     # Used for continuous integration testing with travis
```

I'll mention when you need to add files to `.Rbuildignore` whenever it's important.

### Binary packages

If you want to distribute your package to an R user who doesn't have package development tools, you'll need to make a __binary__ package. Like a package bundle, a binary package is a single file. But if you uncompress it, you'll see that the internal structure is rather different from a source package: 

* There are no `.R` files in the `R/` directory - instead there are three
  files that store the parsed functions in an efficient format. This is
  basically the result of loading all the R code and then saving the
  functions with `save()`. (In the process, this adds a little extra metadata to 
  make things as fast as possible).

* A `Meta/` directory contains a number of `Rds` files. These files contain 
  cached metadata about the package, like what topics the help files cover and
  parsed versions of the `DESCRIPTION` files. (You can use `readRDS()` to see 
  exactly what's in those files). These files make package loading faster
  by caching costly computations.

* An `html/` directory contains files needed for HTML help.

* If you had any code in the `src/` directory there will now be a `libs/`
  directory that contains the results of compiling code for 32 bit
  (`i386/`) and 64 bit (`x64/`).
  
* The contents of `inst/` are moved to the top-level directory.

Binary packages are platform specific: you can't install a Windows binary package on a Mac or vice versa. Also, Mac binary packages end in `.tgz` and Windows binary packages end in `.zip`. You can use `devtools::build(binary = TRUE)` to make a binary package. 

The following diagram summarises the files present in the root directory for the source, bundled and binary versions of devtools.

```{r, echo = FALSE}
bookdown::embed_png("diagrams/package-files.png")
```

### Installed packages {#install}

An __installed__ package is just a binary package that's been decompressed into a package library (described next). The following diagram illustrates the many ways a package can be installed. This diagram is complicated! In an ideal world installing a package would involve stringing together a set of simple steps: source -> bundle, bundle -> binary, binary -> installed. In the real world it's not this simple: each step in the sequence is slow, and there are often faster shortcuts available.

```{r, echo = FALSE}
bookdown::embed_png("diagrams/installation.png")
```

The tool that powers all package installations is the command line tool `R CMD install` - it can install a source, bundle or a binary package. Devtools functions provide wrappers which allow you to access this tool from R rather than from the command line. `install()` is effectively a wrapper for `R CMD install`. `build()` is a wrapper for `R CMD build` that turns source packages into bundles. `install_github()` downloads a source package from GitHub, runs `build()` to make vignettes, and then uses `R CMD install` to do the install. `install_url()`, `install_gitorious()`, `install_bitbucket()` work similarly for packages found elsewhere on the internet.

`install.packages()` and `devtools::install_github()` allow you to install a remote package. They both work by first downloading the package. `install.packages()` normally downloads a binary package built by CRAN. This makes installation very speedy. `install_github()` has to work a little differently - it first downloads a source package, builds it and then installs it.

You can prevent files in the package bundle from being included in the installed package using `.Rinstignore`. This works the same way as `.Rbuildignore`, described above. It is rarely needed.

### In memory packages

To use a package, you must load it into memory. When you're not developing a package, you load it into memory with `library()`. When you are developing a package, you can either use `devtools::load_all()` or RStudio's "Build and reload". You now know enough about packages to understand the difference between the two: `load_all()` skips the installation step and goes directly from on-disk to in-memory.

The following diagram summarises the three ways of loading a package into memory:

```{r, echo = FALSE}
bookdown::embed_png("diagrams/loading.png")
```

## What is a library? {#library}

A library is simply a directory containing installed packages. You can have multiple libraries on your computer and almost every one has at least two: one for packages you’ve installed, and one for the recommended packages that come with every R installation (like base, stats, etc). Normally, that first directory varies based on the version of R that you’re using. That's why it seems like you lose all of your packages when you reinstall R --- they're still on your hard drive, but R can't find them. 

You can use `.libPaths()` to see which libraries are currently active. Here are mine:

```{r, eval = FALSE}
.libPaths()
#> [1] "/Users/hadley/R"                                               
#> [2] "/Library/Frameworks/R.framework/Versions/3.1/Resources/library"
lapply(.libPaths(), dir)
#> [[1]]
#>   [1] "AnnotationDbi"   "ash"             "assertthat"     
#>   ...      
#> [163] "xtable"          "yaml"            "zoo"            
#> 
#> [[2]]
#>  [1] "base"         "boot"         "class"        "cluster"     
#>  [5] "codetools"    "compiler"     "datasets"     "foreign"     
#>  [9] "graphics"     "grDevices"    "grid"         "KernSmooth"  
#> [13] "lattice"      "MASS"         "Matrix"       "methods"     
#> [17] "mgcv"         "nlme"         "nnet"         "parallel"    
#> [21] "rpart"        "spatial"      "splines"      "stats"       
#> [25] "stats4"       "survival"     "tcltk"        "tools"       
#> [29] "translations" "utils"
```

The first lib path is for the packages I've installed (I've installed at lot!). The second is for so-called "recommended" packages that come with every installation of R.

When you use `library(pkg)` to load a package, R looks through each path in `.libPaths()` to see if a directory called `pkg` exists. If it doesn't, you'll get an error message:

```{r, error = TRUE}
library(blah)
```

The main difference between `library()` and `require()` is what happens when a package isn't found. While `library()` throws an error, `require()` prints a message and returns FALSE. In practice this distinction isn't important because when building a package you should __NEVER__ use either inside a package. See [package dependencies](#dependencies) for what you should do instead.

When you start learning R it's easy to get confused between libraries and packages, because you use `library()` function to load a _package_. However, the distinction between libraries and packages is important and useful. One important application is packrat, which automates the process of managing project specific libraries. This means that when you upgrade a package in one project, it only affects that project, not every project on your computer. This is useful because it allows you to play around with cutting-edge packages without affecting other projects' use of older, more reliable packages. This is also useful when you're both developing and using a package. 
