---
layout: default
title: R code
output: bookdown::html_chapter
---

# R code

The most important part of most packages is the `R/` directory because it contains all of your R code. We'll start with `R/`, because even if you do nothing else, putting your R files in this directory gives you some useful tools.

In this chapter you'll learn:

* How to organise the R code in your package.
* Your first package development workflow.
* What happens when you install a package.
* The difference between a library and package.

## Getting started {#getting-started}

The easiest way to get started with a package is to run `devtools::create("path/to/package/pkgname")`. This makes the package directory, `path/to/package/pkgname/`, then adds four things to make the smallest usable package:

1. An RStudio project file, `pkgname.Rproj`.
1. An `R/` directory.
1. A basic `DESCRIPTION` file.
1. A basic `NAMESPACE` file.

In this chapter, you'll learn about the `R/` directory and the RStudio project file. Ignore the other files for now. You'll learn about  the `DESCRIPTION` in [package metadata](#description) and the `NAMESPACE` in [namespaces](#namespace).

__Never__ use `package.skeleton()` to create a package. It's designed for an older era of package development, and mostly serves to make your life harder, not easier. Currently I don't recommend using RStudio's "create a new package" tool because it uses `package.skeleton()`. That will be fixed by the time the book is published.

The first principle of using a package is that all R code goes in `R/`. If you have existing code for your new package, now's a good time to copy it into `R/`.

### Naming tips {#naming}

The hardest thing about creating a new package is coming up with a good name. There's only one formal requirement:

* The package name can only consist of letters and numbers, and must start
  with a letter.

But if you're planning on releasing your package to the wild, I think it's worth a few minutes brainstorming names. I have two recommendations:

* Pick a unique name so you can easily google it. This makes it easy for 
  potential users to find your package, and it makes it easier for you to
  see who's using your package.

* Avoid using both upper and lower case letters: they make the package name
  hard to type and hard to remember. For example, I can never remember if it's
  `Rgtk2` or `RGTK2` or `RGtk2`.

Some strategies I've used in the past to create packages names:

* Find a name evocative of the problem and modify it so that it's unique:
  `plyr` (generalisation of apply tools), `lubridate` (makes dates and times
  easier), `mutatr` (mutable objects), `classifly` (high-dimensional views of
  classification).

* Use abbreviations: `lvplot` (letter value plots), `meifly` (models explored
  interactively).

* Add an extra R: `stringr` (string processing), `tourr` (grand tours), `httr`
  (HTTP requests).

## RStudio projects {#projects}

To get started with your new package in RStudio, double-click the `package.Rproj` file that `create()` just made. This will open a new RStudio project for your package. Projects are a great way to develop packages because:

* Each project is isolated; they keep unrelated things unrelated.

* You get handy code navigation tools like `F2` to jump to a function
  definition and `Ctrl + .` to look up functions by name.

* You get useful keyboard shortcuts for common package development tasks.
  You'll learn these throughout the book, but to see them all press 
  Alt + Shift + K or use the Help | Keyboard shortcuts menu.

`create()` makes an `.Rproj` file for you. If you want to add one to an existing package, use `devtools::use_rstudio("path/to/package")`. If you don't use RStudio, you can get many of the benefits by starting a new R session and ensuring the working directory is set to the project directory. 

`.Rproj` files are just text files. The project file created by devtools looks like this:

```
Version: 1.0

RestoreWorkspace: No
SaveWorkspace: No
AlwaysSaveHistory: Default

EnableCodeIndexing: Yes
Encoding: UTF-8

AutoAppendNewline: Yes
StripTrailingWhitespace: Yes

BuildType: Package
PackageUseDevtools: Yes
PackageInstallArgs: --no-multiarch --with-keep.source
PackageRoxygenize: rd,collate,namespace
```

Never modify this file by hand. Instead, use the friendly project options dialog, accessible from the projects menu in the top-right corner of RStudio.

```{r, echo = FALSE}
bookdown::embed_png("screenshots/project-options-1.png", dpi = 220)
bookdown::embed_png("screenshots/project-options-2.png", dpi = 220)
```

## Organising and running code {#r-code}

The first advantage of using a package is that it's easy to load all of the code in your package. There are two main options: 

* `devtools::load_all()`, __Cmd + Shift + L__, reloads all code in the package. 
  In RStudio, this also saves all open files, saving you a key press.

* Build & reload, __Cmd + Shift + B__. This is only available in RStudio, because
  it installs the package, then restarts R, then loads the package with 
  `library()` (doing this by hand is painful).

These commands support a fluid development workflow:

1. Edit R files in the editor.

1. Press Cmd + Shift + L (or Cmd + Shift + B).

1. Explore the code in the console.

1. Rinse and repeat.

You're free to arrange functions into files however you wish. It's clear that the two extremes are bad: don't put all functions in one file, or every function in its own file. My rule of thumb is that if I can't remember which file a function lives in, I probably need to split them up into more files, or give them better names. It's ok if some files only contain one function, particularly if the function is large or has a lot of documentation. Unfortunately you can't use subdirectories inside in `R/`. The next best thing is to use a common prefix, e.g., `abc-*.R`.

The exact placement of functions within files is less important if you master two important keyboard shortcuts that let you jump to the definition of a function:

* Click a function name in code and press __F2__.

* Press __Ctrl + .__ then start typing the name.

Congratulations, you now understand the basics of using a package! In the rest of this chapter, you'll learn more about the various forms of a package, and exactly what happens when you run `install.packages()` or `install_github()`.

### Avoid side effects

One big difference between a script and R code in a package, is that the code in your package should have no side effects. Code in `R/` should only create objects (mostly functions). Outside of functions, you should not call functions that affect global state:

* Don't use `library()` or `require()` - instead use the 
  [DESCRIPTION](description.html) to say what your package depends on.
  
* Don't modify global `options()` or graphics `par()`. Instead, wrap them
  in a function that the user can call when needed.
  
* Don't save files to disk with `write()`, `write.csv()`, or `saveRDS()`. 
  Instead, use [data/](data.html) to cache important data files.

There are two reasons to avoid side-effects. The first is pragmatic: these funtions will work while you're developing a package locally using `load_all()`, but they won't work once you distribute your package. That's because your R code is run once when the package is built, not everytime `library()` is called. The second is principled: you shouldn't change global state behind your users' backs.

Occassionally, packages do perform side-effects. This is most common if your package talks to an external system - you might need to do some initial setup when the package loads. To do that, you can use the special functions `.onLoad()` and `.onAttach()`. These are called when the package is loaded or attached. You'll learn about the distinction between the two in [Namespaces](#namespace). Unless you know otherwise use `.onLoad()`. Whenever you use `.onLoad()` or `.onAttach()`, make sure to define `.onUnload()` or `.onDetach()` to clean up when the package is unload/detached. 

An older use of `.onAttach()` is to dynamically load a DLL file. This is no longer needed, and should be avoided. See [namespaces](#namespace) for details.  Can still be necessary if you're connecting to external C/C++ libraries that need set up and tear down. 

One type of side-effect that's not so harmful is displaying a message when the package is loaded. This might make clear licensing conditions, or important usage tips. Startup messages should go in `.onAttach()`. To display startup messages, always use `packageStartupMessage()`, and not `message()`. (This allows `suppressPackageStartupMessages()` to selectively suppress package startup messages).

```{r}
.onAttach <- function(...) {
  packageStartupMessage("Welcome to my package")
}
```

## What is a package? {#package}

To develop a packge all you need is the workflow above. But to master package development, particularly distributing your package to others, you need to understand more about the different forms a package can take. This will help you understand exactly what happens when you do `install.packages()`.

So far we've just worked with a __source__ package: the development version of a package that lives on your computer. A source package is just a directory with components like `R/`, `DESCRIPTION`, and so on. There are three other types of package: bundled, binary and installed.

A package __bundle__ is a compressed version of a package in a single file. By convention, package bundles in R use the extension `.tar.gz`. This is Linux convention indicating multiple files have been collapsed into a single file (`.tar`) and then compressed using gzip (`.gz`). A bundle is not that useful in its own right, but is often used an intermediary in other steps. It's rare that you'll need it, but you can call `devtools::build()` to make a package bundle. If you decompress a bundle, you'll see it looks almost the same as your source package. The main differences between a decompressed bundle and a source package are:

* Vignettes, package level documentation, are built turning raw files like
  markdown or latex into output like html and pdf.

* Your source package might contain temporary files used to save time during
  development, like compilation artefacts in `src/`.

* Any files listed in `.Rbuildignore` are not included in the bundle. You'll
  learn more about `.Rbuildignore` in XXX.

If you want to distribute your package to another R user (i.e. someone who doesn't necessarily have the development tools installed) you need to make a __binary__ package. Like a package bundle, a binary package is a single file, but if you uncompress it, you'll see that the internal structure is a little different than a source package: 

* There are no `.R` files in the `R/` directory - instead there are three
  files that store the parsed functions in an efficient format. This is
  basically the result of loading all the R code and then saving the
  functions with `save()`, but with a little extra metadata to make things as
  fast as possible.

* A `Meta/` directory contains a number of `Rds` files. These contain cached
  metadata about the package, like what topics the help files cover and
  parsed versions of the `DESCRIPTION` files. (You can use `readRDS()` to see 
  exactly what's in those files). These files make package loading faster
  by caching costly computations.

* A `html/` directory contains some files needed for HTML help.

* If you had any code in the `src/` directory there will now be a `libs/`
  directory that contains the results of compiling that code for 32 bit
  (`i386/`) and 64 bit (`x64/`).
  
* The contents of `inst/` have been moved into the top-level directory.

Binary packages are platform specific: you can't install a Windows binary package on a Mac or vice versa. Mac binary packages end in `.tgz` and Windows binary packages end in `.zip`. You can use `devtools::build(binary = TRUE)` to make a binary package. 

The following table summarises the files present in the root directory for the source, built and binary versions of devtools.

```{r, results = "asis", echo = FALSE}
knitr::kable(readRDS("extras/pkg-paths.rds"))
```

(Needs diagram instead of table)

* `cran-comments.md` and `devtools.Rproj` are included in `src` but nothing 
  else. That's because they're only needed for development, and are listed in
  `.Rbuildignore`.
  
* `inst/` goes away and `templates/` appears - that's because building a 
  binary copies the contents of `inst/` into the root directory. You'll learn
  more about that in XXX.

### Exercises

1.  Go to CRAN and download the source and binary for XXX. Unzip and compare.
    How do they differ?

1.  Download the __source__ packages for XXX, YYY, ZZZ. What directories do they
    contain?

## Package installation {#install}

An __installed__ package is just a binary package that's been uncompressed into a package library, described next. The following diagram describes the many ways a package can be installed. This diagram is complicated! In an ideal world installing a package would involve stringing together a set of simple steps: source -> bundle, bundle -> binary, binary -> installed, installed -> in memory. It's not this simple in the real world because doing each step in sequence is slow, and there are often faster short cuts available.

```{r, echo = FALSE}
bookdown::embed_png("diagrams/installation.png")
```

The power house is the command line tool `R CMD install` - it can install a source, bundle or a binary package. When you install a package from CRAN, R basically downloads the file and then runs `R CMD install` on it.

Devtools functions wrap the base R functions so that you can access them from R, rather than the command line. `install()` is effectively just a wrapper for `R CMD install`. `build()` is a wrapper for `R CMD build` that turns source packages into bundles. `install_github()` downloads a source package from github, runs `build()` to make vignettes, then uses `R CMD install` to install. `install_url()`, `install_gitorious()`, `install_bitbucket()` work similarly for packages found elsewhere on the internet.

RStudio's "Build and reload" performs one additional step - it also loads the package into memory. `load_all()` is similar, except that it skips all intermediate steps - it goes directly from a source package to a loaded in-memory package. `library()` is the equivalent for installed packages.

## What is a library? {#library}

A collection of packages is called a library. This is a bit confusing because you use the `library()` function to load a package, but the distinction between libraries and packages is important and useful. A library is just a directory containing installed packages. You can have multiple libraries on your computer and almost everyone has at least two: one for the recommended packages that come with a base R install (like `base`, `stats` etc), and one library where the packages you've installed live. Normally, that second directory varies based on the version of R that you're using. That's why it seems like your "lose" all your packages when you reinstall R - they're actually still on your hard drive, but R can't find them. 

You can use `.libPaths()` to see which libraries are currently active. Here are mine:

```{r, eval = FALSE}
.libPaths()
#> [1] "/Users/hadley/R"                                               
#> [2] "/Library/Frameworks/R.framework/Versions/3.1/Resources/library"
lapply(.libPaths(), dir)
#> [[1]]
#>   [1] "AnnotationDbi"   "ash"             "assertthat"     
#>   ...      
#> [163] "xtable"          "yaml"            "zoo"            
#> 
#> [[2]]
#>  [1] "base"         "boot"         "class"        "cluster"     
#>  [5] "codetools"    "compiler"     "datasets"     "foreign"     
#>  [9] "graphics"     "grDevices"    "grid"         "KernSmooth"  
#> [13] "lattice"      "MASS"         "Matrix"       "methods"     
#> [17] "mgcv"         "nlme"         "nnet"         "parallel"    
#> [21] "rpart"        "spatial"      "splines"      "stats"       
#> [25] "stats4"       "survival"     "tcltk"        "tools"       
#> [29] "translations" "utils"
```

My first lib path is where packages that I've installed live (I've installed at lot!), and the second is where packages that come with R live. These are the so called "recommended" packages available with every install of R.

When you use `library(pkg)` to load a package, R looks through each path in `.libPaths()` to see if a directory called `pkg` exists.

Packrat, which we'll learn about in Chapter XXX, automates the process of managing project specific libraries. This means that when you upgrade a package in one project, it only affects that project, not every project on your computer. This is useful because it means you can play around with cutting-edge packages in one place, but all your other projects continue to use the old reliable packages. This is also useful when you're both developing and using a package. 

### Exercises

1.  Where is your default library? What happens when to that library when 
    you install a new package from CRAN?

1.  Can you have multiple version of the same package installed at the same 
    time?

## CRAN notes {#r-cran}

If you're planning on submitting your package to CRAN, you must use only ASCII characters in your `.R` files. You can still include unicode characters in strings, but you need to use the special unicode escape `"\u1234"` format. The easiest way to do that is to use `stringi::stri_escape_unicode()`

```{r}
x <- "This is a bullet •"
y <- "This is a bullet \u2022"
identical(x, y)

cat(stringi::stri_escape_unicode(x))
```

Your R directory should not include any files apart from R code. Subdirectories will be ignored.
