# Software development practices {#sec-sw-dev-practices}

```{r, echo = FALSE}
source("common.R")
status("drafting")
```

## Introduction

In this last part of the book, we zoom back out to consider development practices that can make you more productive and raise the quality of your work.
Here we'll discuss the use of version control and continuous integration.
In @sec-lifecycle we discuss how the nature of package maintenance varies over the lifecycle of a package.

You will notice that we recommend using certain tools:

-   An integrated development environment (IDE).
    In @sec-workflow101-rstudio-projects we encouraged the use of the RStudio IDE for package development work.
    That's what we document, since it's what we use and devtools is developed to work especially well with RStudio.
    But even if it's not RStudio, we strongly recommend working with an IDE that has specific support for R and R package development.

-   Version control.
    We strongly recommend the use of formal version control and, at this point in time, Git is the obvious choice.
    We say that based on Git's general prevalence and, specifically, its popularity within the R package ecosystem.
    In ??,
    we explain why we think version control is so important.

-   Hosted version control.
    We strongly recommend syncing your local Git repositories to a hosted service and, at this point in time, GitHub is the or, at least, "an" obvious choice.
    This the topic of ??.

-   Continuous integration and deployment, a.k.a. CI/CD (or even just CI).
    This terminology comes from the general software engineering world and can sound somewhat grandiose or intimidating when applied to your personal R package.
    All this really means is that you set up specific package development tasks to happen automatically when you push new work to your hosted repository.
    Typically you'll want to run `R CMD check` and to re-build and deploy your package website.
    In ??
    we show how to do this with GitHub Actions.

You might think that these pro-style tools are overkill for someone who doesn't do software development for a living.
While we don't recommend forcing yourself to do everything above on day one of your first "hello world" project, we actually do believe these tools are broadly applicable for R package development.

The main reason is that these tools make it so much easier to do the right thing, e.g. to experiment, document, test, check, and collaborate.
By adopting a shared toolkit, part-time and newer package developers gain access to the same workflows used by experts.
This requires a certain amount of faith and upfront investment, but we believe this pays off.

## Git and GitHub {#sec-sw-dev-practices-git-github}

[Git](https://git-scm.com) is a version control system that was originally built to coordinate the work of a global group of developers working on the Linux kernel.
Git manages the evolution of a set of files --- called a repository --- in a highly structured way and we recommend that every R package should also be a Git repository (and also, probably, an RStudio Project; @sec-workflow101-rstudio-projects).

A solo developer, working on a single computer, will benefit from adopting version control.
But, for most of us, that benefit is not nearly large enough to make up for the pain of installing and using Git.
In our opinion, for most folks, the pros of Git only outweigh the cons once you take the additional step of hooking your local repository up to a remote host like [GitHub](https://github.com).
The joint use of Git and GitHub offers many benefits that more than justify the learning curve.

### Standard practice

This recommendation is well aligned with the current, general practices in software development.
Here are a few relevant facts from the [2022 Stack Overflow developer survey](https://survey.stackoverflow.co/2022/#overview), which is based on about 70K responses.

-   94% report using Git.
    The second-most used version control system was SVN, used by 5% of respondents.

-   For personal projects, 87% of respondents report using GitHub, followed by GitLab (21%) and Bitbucket (11%).
    The ranking is the same albeit less skewed for professional work: GitHub still dominates with 56%, followed by GitLab (29%) and Bitbucket (18%).

We can even learn a bit about the habits of R package developers, based on the URLs found in the `DESCRIPTION` files of CRAN packages.
As of December 2022, there are about 18K packages on CRAN, of which about 55% have a non-empty `URL` field (over 10K).
Of those, 80% have a GitHub URL (over 8K), followed by GitLab (just over 1%) and Bitbucket (around 0.5%).

```{r}
#| eval: false
#| include: false
library(tidyverse)

db <- tools::CRAN_package_db() |> 
  as_tibble() |> 
  select(package = Package, URL)
db

db |> 
  count(has_URL = !is.na(URL)) |> 
  mutate(prop = n / sum(n))
db |> 
  filter(!is.na(URL)) |> 
  mutate(
    github = str_detect(URL, "github"),
    gitlab = str_detect(URL, "gitlab"),
    bitbucket = str_detect(URL, "bitbucket")
  ) |> 
  count(
    github, gitlab, bitbucket
  ) |> 
  mutate(prop = n / sum(n))
```

The prevalence of Git/GitHub, both within the R community and beyond, should help you feel confident that adoption will have tangible benefits.
Furthermore, the sheer popularity of these tools means there are lots of resources available for learning how to use Git and GitHub and for getting unstuck[^software-development-practices-1].

[^software-development-practices-1]: We feature GitHub here, for hosted version control, because it's what we use and what has the best support in devtools.
    However, all the big-picture principles and even some details hold up for alternative platforms, such as Gitlab and Bitbucket.

Two specific resources that address the intersection of Git/GitHub and the R world are the website [Happy Git and GitHub for the useR](https://happygitwithr.com/index.html) and the article "Excuse me, do you have a moment to talk about version control?" [@bryan2018-tas].

We conclude this section with a few examples of why Git/GitHub can be valuable specifically for R package development:

-   Communication with users: GitHub Issues are well-suited for taking bug reports and feature requests.
    Unlike email sent to the maintainer, these conversations are accessible to others and searchable.

-   Collaboration: GitHub pull requests are a very low-friction way for outside contributors to help fix bugs and add features.

-   Distribution: Functions like `devtools::install_github("r-lib/devtools")` and `pak::pkg_install("r-lib/devtools")` allow people to easily install the development version of your package, based on a source repository.
    More generally, anyone can install your package from any valid Git ref, such as a branch, specific SHA, pull request, or tag.

-   Website: GitHub Pages is one of the easiest ways to offer a website for your package (@sec-website-deployment).

-   Continuous integration: This is actually the topic of the next section, so read on for more.

## Continuous integration {#sec-ci}

General intro to CI

## Checking after every commit with GitHub actions {#sec-gha}

If you're using GitHub, as we recommend, you should also use GitHub Actions.
GitHub Actions allow you to run code every time you push to GitHub.

To use GitHub Actions:

1.  Run `usethis::use_github_action_check_standard()` to set up a GitHub Action that runs `R CMD check` on Linux, Mac, and Windows.
    Literally, this adds a new file to your package, below `.github/workflows/.`

2.  Commit this new file and push to GitHub.

3.  Wait a few minutes to see the results in your email.

With this setup in place, `R CMD check` will be run every time you push to GitHub or whenever someone submits a pull request.
You'll find out about failures right away, which makes them easier to fix.
Using automated checks also encourages me to check more often locally, because I know if it fails I'll find out about it a few minutes later, often once I've moved on to a new problem.

### Other uses

Since GitHub Actions allows you to run arbitrary code, there are many other things that you can use it for:

-   Re-publishing a book website every time you make a change to the source.
    (Like this book!)

-   Building vignettes and publishing them to a website.

-   Automatically building a documentation website for your package.

Learn more about using GitHub Actions with R at <https://github.com/r-lib/actions/tree/master/examples>.
