---
layout: default
title: Compiled code
output: bookdown::html_chapter
---

# Compiled code {#src}

* Makefiles beyond the scope of this book. 

* Bundling external libraries beyond the scope of this book (but I wish it 
  wasn't).

* Assumes that you know how to use C or C++ already. If you don't, see the 
  "Advanced R book" (online, amazon).

## C++

Easiest way to access compiled language from R is to use C++.

## C

Sometimes needed if you're wrapping old code, or you really need to eke out performance by avoiding C++ wrappers (this is rare). Mostly important for understand older packages.

*   `.c` files in `src/`, need:

    ```c
    #include <R.h>
    #include <Rdefines.h>
    ```

*  Need wrapper functions like this:

    ```{r, eval = FALSE}
    #' @useDynLib qstack car
    car_ <- function(x) .Call(car, x)
    ```

Speed considerations: <http://cran.r-project.org/doc/manuals/R-exts.html#Speed-considerations>

* Slowest: use `.Call("car")`
* Faster: use `.Call(car)`
* Fastest: don't write wrapper function

This makes microseconds of difference, so normally not important. Only matters if you have tight loop calling millions of times - and you can normally move that loop into C.

If you use `useDynLib`, you also need to add a `.onUnload()` function that cleans up after yourself.

```{r}
.onUnload <- function (libpath) {
  library.dynam.unload("qstack", libpath)
}
```

## Fortran

Fortran 77 subroutines are directly callable via `.Fortran()`, though as with `.C()`, this is inefficient.  The best way to use
Fortran code with R is to write a C wrapper callable by the `.Call()` interface (as described above).  This of course 
requires that the Fortran code can be called by C.  For portability, this generally means either using a non-module
subroutine, or using the Fortran 2003 iso_c_binding module and providing an appropriate interface.  
For more information about Fortran interfaces and mixing C and Fortran, see *Modern Fortran Explained*, by Metcalf, Reid, and Cohen.

## Java

See rJava.  Unlike C and C++ wrappers, must always copy. This usually makes it unacceptably slow when working with larger data (which much of the time is why you're using an other language).

## Dynamically loading dll

Loading - use `@dynlib`.  

To be good member of ecosystem, should also unload:

```{r, eval = FALSE}
.onUnload <- function(libpath) {
  library.dynam.unload("yourpackagename", libpath)
}
```

This ensures that the DLL is unloaded if the package is unloaded.

## Licensing

If you're including code that someone else has written, you need to...

e.g. http://cran.rstudio.com/web/packages/optparse/COPYRIGHTS
