---
layout: default
title: Compiled code
output: bookdown::html_chapter
---

# Compiled code {#src}

It's often useful to include compiled code in an R package. Compiled code, usually  C or C++, is a powerful complement to R code. R code is high-level and expressive, but that expressivity comes at a cost: speed. Low-level compiled languages like C and C++ often require more typing to solve a problem, they can be orders of magnitude faster than R. 

Unfortunately, this book is going to teach you how to write fast C or C++ code. The rest of this chapter will assume you know how to write some C or C++ code. If you don't, feel free to skip this chapter; come back to it later when you need to solve a performance problem.

To learn a compiled language, I recommmend starting with C++ and the Rcpp package. Rcpp makes it very easy to connect C++ to R, and RStudio has a lot of built in tools to facilitate the process. Start by reading my ["High performance functions with Rcpp"](http://adv-r.had.co.nz/Rcpp.html) a freely available book chapter from [Advanced R](http://amzn.com/1466586966?tag=devtools-20): it gives a gentle introduction to C++ by translating familiar R code. Next, check out the [Rcpp book](http://www.rcpp.org/book) and the other resources listed in [learning more](http://adv-r.had.co.nz/Rcpp.html#rcpp-more).

I recommend using `clang` rather than `gcc` to compile C and C++ code. It gives much better error messages. You can make this the default by creating `~/.R/Makevars` which contains

```
CC=clang
CXX=clang++
```

## C++

### Getting started

To get started with Rcpp, run:

```{r, eval = FALSE}
devtools::use_rcpp()
```

This will:

* Create a `src/` directory to hold your `.cpp` files.

* Add `Rcpp` to the `LinkingTo` and `Imports` fields in the `DESCRIPTION`.

* Set up a `.gitignore` file to make sure you don't accidentally check in
  any compiled files (learn more about this in [git](#git)).

* Tells you the two roxygen tags you need to add to your package:

    ```{r}
    #' @useDynLib your-package-name
    #' @importFrom Rcpp sourceCpp
    ```

Once you've done this, the basic workflow is simple:

1.  Create a new C++ file:

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/new-cpp.png", dpi = 220)
    ```

    The default template looks like this:
    
    ```cpp
    #include <Rcpp.h>
    using namespace Rcpp;
    
    // Below is a simple example of exporting a C++ function to R. You can
    // source this function into an R session using the Rcpp::sourceCpp 
    // function (or via the Source button on the editor toolbar)
    
    // For more on using Rcpp click the Help button on the editor toolbar
    
    // [[Rcpp::export]]
    int timesTwo(int x) {
       return x * 2;
    }
    ```
    
    It includes a basic function and some instructions to get started.

1.  Press Cmd + Shift + L (or Cmd + Shift + B).

1.  Run `timesTwo(10)` to confirm that the code works.

When you re-load or re-build the package, devtools runs `Rcpp::compileAttributes()`. This inspects your `.cpp` functions looking for __attributes__ of the form `// [[Rcpp::export]]`. When it finds one, it generates the code necessary to make the function available in R, and creates `src/RcppExports.cpp` and `R/RcppExports.R`. You should never modify these files by hand.

### Documentation

You can use roxygen2 documentation inside C++ files: 

```cpp
//' Multiply a number by two
//' 
//' @param x A single integer.
//' @export
// [[Rcpp::export]]
int timesTwo(int x) {
   return x * 2;
}
```

The two exports mean slightly different things:

* `[[Rcpp::export]]`: make the C++ function available to R.
* `@export`: make the R wrapper function available outside your package.

### Sharing with other packages

### Best practices

* To print output use `Rcout << ...` not `cout << ...`. This always prints to 
  the right place, which might be a GUI console or a file (if `sink()` 
  is active)

* In long-running loops, reguarly `Rcpp::checkUserInterrupt()`. This checks 
  for user interrupts, so a long-running process can be gracefully exited.

* Read Martyn Plummer's article on [Portable C++ for R packages](http://journal.r-project.org/archive/2011-2/RJournal_2011-2_Plummer.pdf)

* Whenever you use C++ code in your package, you should be a good citizen 
  and clean up after yourself if your package is unloaded. This basically 
  means that you should have a `.onUnload()` function that unloads the DLL 
  that Rcpp dynmically loaded for you.

    ```{r}
    .onUnload <- function (libpath) {
      library.dynam.unload("mypackage", libpath)
    }
    ```


## C

If you're writing new compiled code, it's almost always better to use C++. It's more consistent, better documented, and has better tooling. There are three reasons you might need to use C:

* You're working with an older package that uses the C API.
* You're binding to an existing C library.
* You want to eke out every last drop of performance.

### Getting started

There are two sides to calling a C function from R. First you write the C function. It goes in `.c` file in `src/`. Every C file that works with R data structures must include two header files:

```c
#include <R.h>
#include <Rinternals.h>
```

(Including `<Rinternals.h>` seems like bad form, but it doesn't actually give you access to the "internal" internal API unless you set some additional flags. The default just gives you the "external" internal API which is both safe and necessary. Yes, this is confusing.)

These headers allow you to access R's C API. Unfortunately this API is not well documented. I'd recommend starting with my notes at [R's C interface](http://adv-r.had.co.nz/C-interface.html). After that, read [R extensions](http://cran.rstudio.com/doc/manuals/r-devel/R-exts.html). Since many of the API functions are not documented, you'll often also need to read the [R source code](https://github.com/wch/r-source) to figure out the details.

The absolute minimum that you need to know is that C functions which interface to R code must use the `SEXP` type for both inputs and outputs. `SEXP`, short for s-expression, is the C type used to represent every object in R. Every C function must start with a series of conversions from R types to C types, and often ends by converting C types back to R types. (The R API is designed so that these conversions often don't require allocating new memory.) For example to convert length one R vectors back and forth between C scalars, you use the following functions:

R type      | C type      | R -> C            |  C -> R 
------------|-------------|-------------------|-------------------
integer     | int         | `asInteger(x)`    | `ScalarInteger(x)`
numeric     | double      | `asReal(x)`       | `ScalarReal(x)`
logical     | int         | `asLogical(x)`    | `ScalarLogical(x)`
chararacter | const char* | `CHAR(asChar(x))` | `mkString(x)`

We now have enough information to write a simple C function to add two numbers together:

```c
#include <R.h>
#include <Rinternals.h>

SEXP add_(SEXP x_, SEXP y_) {
  double x = asReal(x_);
  double y = asReal(y_);
  
  double sum = x + y;
  
  return ScalarReal(sum);
}
```

To access this C function from R, we need a wrapper that calls `.Call()`:

```{r}
#' @useDynLib mypackage add_
add <- function(x, y) .Call(add_, x, y)
```

There are two ways to call C functions from R: `.C()` and `.Call()`. `.C()` is an older interface and should not be used for new code. You'll need to read about it yourself in [Interface .C and .Fortran functions](http://cran.r-project.org/doc/manuals/R-exts.html#Interface-functions-_002eC-and-_002eFortran) in R-exts. All modern code should use `.Call()`.

Where does the first argument to `.Call()`, `add_` come from? The `@useDynLib` directive is important - it creates a line in the NAMESPACE that looks like this:

```
useDynLib(mypackage, add_)
```

This creates an object called `add_` in the package environment. `add_` is a C function pointer with a little additional metadata:

```{r, eval = FALSE}
add
#> $name
#> [1] "add_"
#> 
#> $address
#> <pointer: 0x107be3f40>
#> $package
#> NULL
#> 
#> attr(,"class")
#> [1] "NativeSymbolInfo"
```

`.Call()` takes pointer to a C function and calls it. It does not check the arguments (either number or type), so you need to do that in your wrapper function, in your C function, or live with R crashing any time you accidentally supply the wrong type of input.

### Sharing with other packages

If you want to make your C functions available to other packages (without going through the R wrapper), you need to "register" the functions. You do this by creating an array of functions to register (containing the name, a function pointer, and the number of arguments) and then calling `R_registerRoutines()` in the initialisation function (`R_init_mypackage()`, called automatically when the "mypackage" DLL is loaded). For example, to register the `add_` function defined above, you'd write C like:

```c
#include "add.h"
#include <R_ext/Rdynload.h>

R_CallMethodDef callMethods[]  = {
  {"add_", (DL_FUNC) &add_, 2},
  {NULL,   NULL,            0}
};

void R_init_mypackage(DllInfo *info) {
  R_registerRoutines(info, NULL, callMethods, NULL, NULL);
}
```

Note the use of `#include add.h`: you'll also need to make a header file that defines the external interface to your C files.

Typically this code is put in a file called `init.c`. The other arguments to `R_registerRoutines` are used for more esoteric interfaces: `.C`, `.Fortran` and `.External`. For more details, see [Registering native extensions](http://cran.r-project.org/doc/manuals/R-exts.html#Registering-native-routines) in R-exts.

If you choose to use registration, you don't need a `@useDynLib` directive for each C function. Instead you can make all symbols available with a single tag: `@useDynLib mypackage, .registration = TRUE`.

### Best practices

* Avoid calls to `assert()`, `abort()` and `exit()`: these will kill the 
  R process, not just your C/C++ code.

* To print output use `Rprintf()`, not `printf()`. This always prints to 
  the write place, which might be a GUI console, or a file (if `sink()` 
  is active)

* In long-running loops, reguarly call `R_CheckUserInterrupt()` to allow
  the user to interrupt C code.

* Use R's macros `ISNAN` and `R_FINITE` to check for NaNs and infinite
  values - these work on more platforms than the C99 `isnan` and `isfinite`.

* Like for C++, whenever you use C code in your package, you should unload the 
  DLL when the package is unloaded:

    ```{r}
    .onUnload <- function (libpath) {
      library.dynam.unload("mypackage", libpath)
    }
    ```

## Makefiles

Makefiles are beyond the scope of this book, but if you're interesting in learning about them, start with the reproducible research motivation at <http://kbroman.org/minimal_make/>. Generally, R packages should avoid a custom `Makefile` in an R package. 

Instead of a full `Makefile`, you should use `Makevars`. `Makevars` is a makefile, but it's included with the default makefile generated by R. `Makevars` allows you to take advantage of R's default behaviour, while being able to selectively set specific flags for your package:

* `PKG_CFLAGS` & `PKG_CXXFLAGS`: C and C++ flags. Most commonly used to set 
  define directives with `-D`.

* `PKG_CPPFLAGS`: Pre-processor flags (not C++ flags!). Most commonly used to
  set include directories with `-I`. Any package listed in the `LinkingTo` field
  in the `DESCRIPTION` will be automatically included.

* `PKG_LIBS`: Linker flags.

`Makevars` must work across many platforms, so don't use gcc extensions. (Or just require GNU make by stating `SystemRequirements: GNU make`)

* Use LF for line breaks, and end with a trailing LF.

* Do not use GNU extensions like conditional inclusions (ifeq), `${shell ...}` 
  (use backticks instead), `${wildcard ...}`, `+=` or `:=`. 

* Avoid bash extensions in expressions passed to the shell.

* Only set portable flags. In particular, note that `-Wall`, `-pedantic`
  and `-O0` are not portable.

To set flags only on Windows, use `Makevars.win`. To build a `Makevars` with `configure`, use `Makevars.in`.

## Other languages

* __Fortran__: It's possible to call Fortran 77 subroutines directly with 
  `.Fortran()`. See the documentation for details. Note that in one package you
  can mingle C and C++, and C and Fortran, but not C++ and Fortran.

* __Java__: The [rJava](https://github.com/s-u/rJava) package makes it 
  possible to call Java code from within R. Note that unlike for C and C++,
  passing an R object to a java call must make a copy, which has serious
  performance implications.

## Licensing

If your package includes code written by someone else you must ensure that the package license is compatible with the license for the included code:

* The simplest solution is to use the same license for your code and the 
  included code. 

* If you don't want to use the same license, you're best sticking
  with common cases where the interactions are well known. For example, 
  [Various Licenses and Comments about Them](https://www.gnu.org/licenses/license-list.html)
  describes what licenses are compatible with the GPL license.
  
  In this case your description should contain 
  `License: <main license> + FILE license` where main license is a license valid
  for the complete bunlde of code, and the `license` file describes the 
  licenses of the individual components.

* For non-standard cases, you'll need to consult a lawyer.

In all cases, make sure you include copyright and license statements from the original code.

## CRAN issues

You can not include any binary executables in your package. Instead, the package must build all binary artefact from source all major platforms (linux, windows and mac). To opt-out of this requirement, you will need a compelling reason, like you're binding to functionality that's only available on one OS.

R-exts has some advice on [writing portable C and C++ code](http://cran.rstudio.com/doc/manuals/r-devel/R-exts.html#Portable-C-and-C_002b_002b-code).

The interface between CRAN's automated and manual checking can be particularly frustrating when your package contains compiled code. Requirements vary from submission to submission, based on which maintainer you get and how much free time they have. The rules are inconsistently applied, but if they happen to you, it's best to bite the bullet and make the change rather than trying to argue about it:

* Sometimes you will need to list all authors and copyright holders of included 
  code in the DESCRIPTION. 

* Sometimes your package will need to work on Solaris. Fixing Solaris issues 
  can be hard, due to the difficulty of accessing a computer running Solaris. 
  You will be in a stronger negotiating position if the package has no problems 
  on other platforms. 
  

