---
layout: default
title: Compiled code
output: bookdown::html_chapter
---

# Compiled code {#src}

It's often useful to include compiled code in an R package. Compiled code, whether it's C, C++ or Fortran, is a powerful complement to R code. R code is high-level and expressive, but that expressivity comes at a cost: speed. Low-level languages like C and C++ often require more typing to solve a problem, they can be multiple orders of magnitude faster than R. 

Teaching you how to write C and C++ code is beyond the scope of the book. If you don't already know a compiled language, I'd recommmend learing C++ and the Rcpp package. The Rcpp package makes it incredibly easy to connect C++ to R, and RStudio has a lot of built in tools to facilitate the process. If you'd like to learn how to write Rcpp code, I'd recommend starting with ["High performance functions with Rcpp"](http://adv-r.had.co.nz/Rcpp.html) a freely available book chapter from [Advanced R](http://amzn.com/1466586966?tag=devtools-20). Once you've done that, check out the [Rcpp book](http://www.rcpp.org/book) and the other resources listed in [learning more](http://adv-r.had.co.nz/Rcpp.html#rcpp-more).

The rest of this chapter will assume you know how to write some C or C++ code. If you don't, feel free to skip this chapter; come back to it later when you need to solve a performance problem.

## C++

Easiest way to access compiled language from R is to use C++. To get started, run:

```{r, eval = FALSE}
devtools::use_rcpp()
```

This will:

* Create a `src/` directory where you'll put your `.cpp` files.

* Adds `Rcpp` to the `LinkingTo` and `Imports` fields in the `DESCRIPTION`.

* Set up a `.gitignore` file to make sure you don't accidentally check in
  any compiled files (if you're using [git](#git)).

* Tells you about two roxygen tags you need to add to your package:

    ```{r}
    #' @useDynLib your-package-name
    #' @importFrom Rcpp sourceCpp
    ```

### Workflow

Once you've done this, the basic workflow is simple:

1. Create a new C++ file:

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/new-cpp.png", dpi = 220)
    ```

1. Press Cmd + Shift + L (or Cmd + Shift + B).

1. Explore the code in the console.

1. Edit the file and repeat the previous steps.

If you have any compile errors, they will appear in a special pane at the top-left.

## C

If you're writing new compiled code, it's almost always better to use Rcpp. There are three reasons to understand R's C API:

* You're working with a package written before Rcpp that uses the C API.
* You're binding to another C library.
* You want to eke out every last drop of performance.

There are two sides to accessing a C function in R. First you write the C function. It goes in `.c` file in `src/`. Every C file that interfaces with R must contain two header files:

```c
#include <R.h>
#include <Rdefines.h>
```

These allow you to access R's C API. Unfortunately this API is not terribly well documented anywhere. I'd recommend starting with [R's C interface](http://adv-r.had.co.nz/C-interface.html). After that you'll need to read [R extensions](http://cran.rstudio.com/doc/manuals/r-devel/R-exts.html) and a lot of [R's source code](https://github.com/wch/r-source).

The absolute minimum that you need to know is that C functions that talk to R code must use the `SEXP` for both inputs and outputs. At the heart of every C function is a set of conversions between R objects and C objects. For example to convert length one vectors back and forth between C scalars, you use the following functions:

R type      | C type      | R -> C            |  C -> R 
------------|-------------|-------------------|-------------------
integer     | int         | `asInteger(x)`    | `ScalarInteger(x)`
numeric     | double      | `asReal(x)`       | `ScalarReal(x)`
logical     | int         | `asLogical(x)`    | `ScalarLogical(x)`
chararacter | const char* | `CHAR(asChar(x))` | `mkString(x)`

A simple C function to add two numbers together would look like this:

```c
#include <R.h>
#include <Rdefines.h>

SEXP add_(SEXP x_, SEXP y_) {
  double x = asReal(x_);
  double y = asReal(y_);
  
  double sum = x + y;
  
  return ScalarReal(sum);
}
```

(Working with vectors is considerably more complicated.)

As well as this C function, you also need an R wrapper:

```{r}
#' @useDynLib mypackage add_
add <- function(x, y) .Call(add_, x, y)
```

The `@useDynLib` directive is important - it creates a line in the NAMESPACE that looks like

```
useDynLib(mypackage, add_)
```

This will create an object called `add_` in the package environment. This object represents a pointer to the C function in memory.

```{r, eval = FALSE}
add
#> $name
#> [1] "add_"
#> 
#> $address
#> <pointer: 0x107be3f40>
#> $package
#> NULL
#> 
#> attr(,"class")
#> [1] "NativeSymbolInfo"
```

`.Call()` takes pointer to a C function and calls it. It does not checking of the arguments (either number of type), so you'll either have to do that checking in your wrapper function or in your C function, or live with R crashing any time you accidentally supply the wrong type of argument.

R also provides tools for C function "registration". But it doesn't seem to be worth bothering - it's a lot double, adds duplication to your code and if anything, is a little slower.

Whenever you use C or C++ code in your package, you should be a good citizen and clean up after yourself if your package is unloaded. This basically means that you should have a `.onUnload()` function that unloads the DLL that `useDynLib` dynmically loaded for you.

```{r}
.onUnload <- function (libpath) {
  library.dynam.unload("mypackage", libpath)
}
```

## Other languages

### Fortran

Fortran 77 subroutines are directly callable via `.Fortran()`, though as with `.C()`, this is inefficient.  The best way to use Fortran code with R is to write a C wrapper callable by the `.Call()` interface (as described above).  This of course requires that the Fortran code can be called by C.  For portability, this generally means either using a non-module subroutine, or using the Fortran 2003 `iso_c_binding` module and providing an appropriate interface. For more information about Fortran interfaces and mixing C and Fortran, see *Modern Fortran Explained*, by Metcalf, Reid, and Cohen.

Note that in one package you can mingle C and C++, and C and Fortran, but not C++ and Fortran.

### Java

See rJava.  Unlike C and C++ wrappers, must always copy. This usually makes it slow when working with larger data (which much of the time is why you're using an other language).

## Dynamically loading dll

To import compiled code from another package, use `@useDynLib`

* `@useDynLib package` imports all compiled functions.

* `@useDynLib package routinea routineb` imports selected compiled functions.

* Any `@useDynLib` specification containing a comma, e.g.
  `@useDynLib mypackage, .registration = TRUE` will be inserted as is
  into the the `NAMESPACE`, e.g. `useDynLib(mypackage, .registration = TRUE)`


To be good member of ecosystem, should also unload:

```{r, eval = FALSE}
.onUnload <- function(libpath) {
  library.dynam.unload("yourpackagename", libpath)
}
```

This ensures that the DLL is unloaded if the package is unloaded.

## Licensing

If you're including code that someone else has written, you need to...

e.g. http://cran.rstudio.com/web/packages/optparse/COPYRIGHTS

## Makefiles

Makefiles beyond the scope of this book, but see <http://kbroman.org/minimal_make/> to get started with a reproducible research example.

Avoid in R packages. Instead use a Makevars to set options. Sourced in before the standard R makefile.

> The most common use of a Makevars file is to set additional preprocessor options (for example include paths) for C/C++ files via PKG_CPPFLAGS, and additional compiler flags by setting PKG_CFLAGS, PKG_CXXFLAGS, PKG_FFLAGS or PKG_FCFLAGS, for C, C++, FORTRAN or Fortran 9x respectively"

> N.B.: Include paths are preprocessor options, not compiler options, and must be set in PKG_CPPFLAGS as otherwise platform-specific paths (e.g. ‘-I/usr/local/include’) will take precedence.

## CRAN issues

Solaris.

> Package authors should make all reasonable efforts to provide cross-platform portable code. Packages will not normally be accepted that do not run on at least two of the major R platforms. Cases for Windows-only packages will be considered, but CRAN may not be the most appropriate place to host them.
