---
layout: default
title: Compiled code
output: bookdown::html_chapter
---

# Compiled code {#src}

It's often useful to include compiled code in an R package. Compiled code, usually  C or C++, is a powerful complement to R code. R code is high-level and expressive, but that expressivity comes at a cost: speed. Low-level compiled languages like C and C++ often require more typing (and more thinking) to solve a problem, they can be orders of magnitude faster than R. 

Unfortunately, teaching you how to program in C or C++ is beyond the scope of the book.
If you'd like to learn, I recommmend starting with C++ and the Rcpp package. Rcpp makes it very easy to connect C++ to R, and RStudio has many tools to facilitate the process. Start by reading my ["High performance functions with Rcpp"](http://adv-r.had.co.nz/Rcpp.html), a freely available book chapter from [Advanced R](http://amzn.com/1466586966?tag=devtools-20): it gently introduces you to C++ by translating familiar R code. Next, check out the [Rcpp book](http://www.rcpp.org/book) and the other resources listed in [learning more](http://adv-r.had.co.nz/Rcpp.html#rcpp-more).

## C++

### Getting started

To get your package setup to use Rcpp, run:

```{r, eval = FALSE}
devtools::use_rcpp()
```

This will:

* Create a `src/` directory to hold your `.cpp` files.

* Add `Rcpp` to the `LinkingTo` and `Imports` fields in the `DESCRIPTION`.

* Set up a `.gitignore` file to make sure you don't accidentally check in
  any compiled files (learn more about this in [git](#git)).

* Tells you the two roxygen tags you need to add to your package:

    ```{r}
    #' @useDynLib your-package-name
    #' @importFrom Rcpp sourceCpp
    NULL
    ```

### Workflow

Once you're set up, the basic workflow is familiar:

1.  Create a new C++ file:

    ```{r, echo = FALSE}
    bookdown::embed_png("screenshots/new-cpp.png", dpi = 220)
    ```

    The default template looks like this:
    
    ```cpp
    #include <Rcpp.h>
    using namespace Rcpp;
    
    // Below is a simple example of exporting a C++ function to R. You can
    // source this function into an R session using the Rcpp::sourceCpp 
    // function (or via the Source button on the editor toolbar)
    
    // For more on using Rcpp click the Help button on the editor toolbar
    
    // [[Rcpp::export]]
    int timesTwo(int x) {
       return x * 2;
    }
    ```
    
    It includes a basic function and some instructions to get started.

1.  Generate the necessary modifications to your `NAMESPACE`, by documenting
    with Cmd + Shift + D.

1.  Press Cmd + Shift + L to load all code (or Cmd + Shift + B to
    build and reload).

1.  Run `timesTwo(10)` from the console to check that it works.

Behind the scenes, `devtools::load_all()` and RStudio's "Build and reload" do a lot of work for you. They:

* Set up your R environment to compile code and warn you if you're missing 
  necessary pieces.
  
* Call `Rcpp::compileAttributes()`. This inspects your `.cpp` functions 
  looking for __attributes__ of the form `// [[Rcpp::export]]`. When it finds 
  one, it generates the code necessary to make the function available in R, 
  and creates `src/RcppExports.cpp` and `R/RcppExports.R`. You should never 
  modify these files by hand.
  
* Build a DLL (dynamically linked library) and make it available to R.

### Documentation

Each C++ function that is exported to R comes with an R wrapper function. For example, the R `timesTwo()` function looks like:

```R
timesTwo <- function(x) {
  .Call('timesTwo', PACKAGE = 'mypackage', x)
}
```

You can use roxygen2 to document this like a regular R function, but instead of using `#'` you use `\\'` (the C++ commenting convention):

```cpp
//' Multiply a number by two
//' 
//' @param x A single integer.
//' @export
// [[Rcpp::export]]
int timesTwo(int x) {
   return x * 2;
}
```

The distinctions between the two export directives is important:

* `[[Rcpp::export]]`: make the C++ function available to R.
* `@export`: make the R wrapper function available outside your package.

### Sharing with other packages

To make you C++ code callable by other packages, add:

```cpp
// [[Rcpp::interfaces(r, cpp)]]
```

Rcpp takes care of the details of generating a header file in the right place. See [Rcpp Attributes](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-attributes.pdf) for more details, including how to combine hand written and automatically generated header files.

### Using C++ from other packages

Follow these three steps to use Rcpp C++ code from another package. For example, to use C++ code from otherPackage:

1. Add `LinkingTo: otherPackage` to the `DESCRIPTION`.

1. In the C++ file that uses code from otherPackage add:

    ```cpp
    // [[Rcpp::depends(otherPackage)]]
    #include <otherPackage.h>
    ```

1. Functions from otherPackage will be included in the `otherPackage` C++
   namespace. Use `otherPackage::foo()` to access functions.

### Best practices

* To print output use `Rcout << ...` (not `cout << ...`). This always prints to 
  the right place, which might be a GUI console or a file (if `sink()` 
  is active)
  
* Throw errors with `stop()`, not `throw()`.

* In long-running loops, regularly run `Rcpp::checkUserInterrupt()`. This 
  checks for user interrupts, and aborts your C++ if the user has pressed
  Ctrl + C or Escape in R.

* Follow Martyn Plummer's recommendations on 
  [Portable C++ for R packages](http://journal.r-project.org/archive/2011-2/RJournal_2011-2_Plummer.pdf).

* Whenever you use C++ code in your package, you need to clean up after 
  yourself when your package is unloaded. Do this by writing a `.onUnload()` 
  function that unloads the DLL:

    ```{r}
    .onUnload <- function (libpath) {
      library.dynam.unload("mypackage", libpath)
    }
    ```

* Use `clang` instead of `gcc` to compile your C++ code: it gives much 
  better error messages. You can make `clang` the default by creating a
  `~/.R/Makevars` that contains:

    ```bash
    CXX=clang++
    ```

## C

If you're writing new compiled code, it's almost always better to use C++. It's more consistent, better documented, and has better tooling. However, there are some reasons you might choose to use C:

* You're working with an older package that already uses the C API.
* You're binding to an existing C library.
* You want to eke out every last drop of performance.

### Getting started

There are two sides to calling a C function from R. First you write the C function. It goes in a `.c` file in `src/`. Every C file that works with R data structures must include two header files:

```c
#include <R.h>
#include <Rinternals.h>
```

(Including `<Rinternals.h>` seems like bad form, but it doesn't actually give you access to the "internal" internal API unless you set some additional flags. The default just gives you the "public" internal API which is both safe and necessary. Yes, this is confusing.)

These headers allow you to access R's C API. Unfortunately this API is not well documented. I'd recommend starting with my notes at [R's C interface](http://adv-r.had.co.nz/C-interface.html). After that, read [The R API](http://cran.rstudio.com/doc/manuals/r-devel/R-exts.html#The-R-API) in R-exsts. Since many of the API functions are not documented, you'll often also need to read the [R source code](https://github.com/wch/r-source) to figure out the details.

Here's the minimum you need to know: C functions which talk to R must use the `SEXP` type for both inputs and outputs. `SEXP`, short for s-expression, is the C type used to represent every type of object in R. A C function typically starts by converting `SEXP`s to C data types, and ends by converting back to a `SEXP`. (The R API is designed so that these conversions often don't require allocating new memory.) For example, to convert length one R vectors back and forth between C scalars, you use the following functions from the R API:

R type      | C type      | R -> C            |  C -> R 
------------|-------------|-------------------|-------------------
integer     | int         | `asInteger(x)`    | `ScalarInteger(x)`
numeric     | double      | `asReal(x)`       | `ScalarReal(x)`
logical     | int         | `asLogical(x)`    | `ScalarLogical(x)`
chararacter | const char* | `CHAR(asChar(x))` | `mkString(x)`

We now have enough information to write a simple C function that can add two numbers together:

```c
#include <R.h>
#include <Rinternals.h>

SEXP add_(SEXP x_, SEXP y_) {
  double x = asReal(x_);
  double y = asReal(y_);
  
  double sum = x + y;
  
  return ScalarReal(sum);
}
```

To access this C function from R, we need a wrapper that calls `.Call()`:

```{r}
#' @useDynLib mypackage add_
add <- function(x, y) .Call(add_, x, y)
```

There are two ways to call C functions from R: `.C()` and `.Call()`. `.C()` is an older interface and should not be used for new code. You'll need to read about it yourself in [Interface .C and .Fortran functions](http://cran.r-project.org/doc/manuals/R-exts.html#Interface-functions-_002eC-and-_002eFortran) in R-exts. All modern code should use `.Call()`.

Where does the first argument to `.Call()`, `add_` come from? The `@useDynLib` directive is important - it creates a line in the NAMESPACE that looks like this:

```
useDynLib(mypackage, add_)
```

This creates an object called `add_` in the package environment. `add_` is a C function pointer with a little additional metadata:

```{r, eval = FALSE}
add
#> $name
#> [1] "add_"
#> 
#> $address
#> <pointer: 0x107be3f40>
#> $package
#> NULL
#> 
#> attr(,"class")
#> [1] "NativeSymbolInfo"
```

`.Call()` takes pointer to a C function and calls it. It does not check the arguments (either number or type), so you need to do that in your wrapper function, in your C function, or live with R crashing any time you accidentally supply the wrong type of input.

### Sharing with other packages

There are two components to making your C code available in other packages: the header file and function registration. 

Start by creating a header file that describes the external interface to your code. It should live in `inst/includes/mypackage.h`. You'll need to add the following environment variables to your `src/Makevars` to make sure that your code can also find those headers:

```{r}
PKG_CXXFLAGS=-I../inst/include
```

The DLLs in R packages must be relocatable, because most R users get binaries from CRAN, and they are installed in many different places. To allow you to call a C function in an arbtirary location, R provides a function registration API. To register a function you call `R_RegisterCCallable()` from `R_init_mypackage()` which is called automatically when the "mypackage" DLL is loaded. Typically this code is put in a file called `init.c`. 

```c
#include "add.h"
#include <R_ext/Rdynload.h>

void R_init_mypackage(DllInfo *info) {
  R_RegisterCCallable(info, "add_",  (DL_FUNC) &add_)
}
```

Confusingly, there's another type of function registration. Instead of using `@useDynLib pkg fun` you can make functions available to R en masse by calling a C api and then doing `@useDynLib mypackage, .registration = TRUE`. Do so by in C creating an array of functions to register (containing the name, a function pointer, and the number of arguments) and then calling `R_registerRoutines()` in the initialisation function. For example, to register the `add_` function defined above, you'd write C like:

```c
#include "add.h"
#include <R_ext/Rdynload.h>

R_CallMethodDef callMethods[]  = {
  {"add_", (DL_FUNC) &add_, 2},
  {NULL,   NULL,            0}
};

void R_init_mypackage(DllInfo *info) {
  R_registerRoutines(info, NULL, callMethods, NULL, NULL);
}
```

This is only useful if useful if you're using the older `.C` interface, or the more esoteric `.Fortran` and `.External`. The other arguments to `R_registerRoutines` are used for more esoteric interfaces: `.C`, . For more details, see [Registering native extensions](http://cran.r-project.org/doc/manuals/R-exts.html#Registering-native-routines) in R-exts. 

### Using C code from another package

`R_GetCCallable`

### Best practices

* Avoid calls to `assert()`, `abort()` and `exit()`: these will kill the 
  R process, not just your C/C++ code.

* To print output use `Rprintf()`, not `printf()`. This always prints to 
  the write place, which might be a GUI console, or a file (if `sink()` 
  is active)

* In long-running loops, reguarly call `R_CheckUserInterrupt()` to allow
  the user to interrupt C code.

* Use R's macros `ISNAN` and `R_FINITE` to check for NaNs and infinite
  values - these work on more platforms than the C99 `isnan` and `isfinite`.

* Like for C++, whenever you use C code in your package, you should unload the 
  DLL when the package is unloaded:

    ```{r}
    .onUnload <- function (libpath) {
      library.dynam.unload("mypackage", libpath)
    }
    ```

* Use `clang` instead of `gcc` to compile your C code: it gives much 
  better error messages. You can make `clang` the default by creating a
  `~/.R/Makevars` that contains:

    ```bash
    C=clang++
    ```

## Makefiles

Makefiles are beyond the scope of this book, but if you're interesting in learning about them, start with the reproducible research motivation at <http://kbroman.org/minimal_make/>. Generally, R packages should avoid a custom `Makefile` in an R package. 

Instead of a full `Makefile`, you should use `Makevars`. `Makevars` is a makefile, but it's included with the default makefile generated by R. `Makevars` allows you to take advantage of R's default behaviour, while being able to selectively set specific flags for your package:

* `PKG_CFLAGS` & `PKG_CXXFLAGS`: C and C++ flags. Most commonly used to set 
  define directives with `-D`.

* `PKG_CPPFLAGS`: Pre-processor flags (not C++ flags!). Most commonly used to
  set include directories with `-I`. Any package listed in the `LinkingTo` field
  in the `DESCRIPTION` will be automatically included.

* `PKG_LIBS`: Linker flags.

`Makevars` must work across many platforms, so don't use gcc extensions. (Or just require GNU make by stating `SystemRequirements: GNU make`)

* Use LF for line breaks, and end with a trailing LF.

* Do not use GNU extensions like conditional inclusions (ifeq), `${shell ...}` 
  (use backticks instead), `${wildcard ...}`, `+=` or `:=`. 

* Avoid bash extensions in expressions passed to the shell.

* Only set portable flags. In particular, note that `-Wall`, `-pedantic`
  and `-O0` are not portable.

To set flags only on Windows, use `Makevars.win`. To build a `Makevars` with `configure`, use `Makevars.in`.

## Other languages

* __Fortran__: It's possible to call Fortran 77 subroutines directly with 
  `.Fortran()`. See the documentation for details. Note that in one package you
  can mingle C and C++, and C and Fortran, but not C++ and Fortran.

* __Java__: The [rJava](https://github.com/s-u/rJava) package makes it 
  possible to call Java code from within R. Note that unlike for C and C++,
  passing an R object to a java call must make a copy, which has serious
  performance implications.

## Licensing

If your package includes code written by someone else you must ensure that the package license is compatible with the license for the included code:

* The simplest solution is to use the same license for your code and the 
  included code. 

* If you don't want to use the same license, you're best sticking
  with common cases where the interactions are well known. For example, 
  [Various Licenses and Comments about Them](https://www.gnu.org/licenses/license-list.html)
  describes what licenses are compatible with the GPL license.
  
  In this case your description should contain 
  `License: <main license> + FILE license` where main license is a license valid
  for the complete bunlde of code, and the `license` file describes the 
  licenses of the individual components.

* For non-standard cases, you'll need to consult a lawyer.

In all cases, make sure you include copyright and license statements from the original code.

## CRAN issues

You can not include any binary executables in your package. Instead, the package must build all binary artefact from source all major platforms (linux, windows and mac). To opt-out of this requirement, you will need a compelling reason, like you're binding to functionality that's only available on one OS.

R-exts has some advice on [writing portable C and C++ code](http://cran.rstudio.com/doc/manuals/r-devel/R-exts.html#Portable-C-and-C_002b_002b-code).

The interface between CRAN's automated and manual checking can be particularly frustrating when your package contains compiled code. Requirements vary from submission to submission, based on which maintainer you get and how much free time they have. The rules are inconsistently applied, but if they happen to you, it's best to bite the bullet and make the change rather than trying to argue about it:

* Sometimes you will need to list all authors and copyright holders of included 
  code in the DESCRIPTION. 

* Sometimes your package will need to work on Solaris. Fixing Solaris issues 
  can be hard, due to the difficulty of accessing a computer running Solaris. 
  You will be in a stronger negotiating position if the package has no problems 
  on other platforms. 
  

